---
title: "Preparation of MetaPhlAn dataset and transformation of data"
output:
  github_document:
    toc: true
editor_options:
  chunk_output_type: console
---
# Load in the packages and path to folder:
```{r echo=T, message=F, error=F, warning=F}
library(phyloseq)
library(vegan)
library(zCompositions)
library(rgr)
library(NBZIMM)
library(dplyr)
library(ggplot2)
library(readr)
library(ggpubr)
path = "C:/Users/FX76TZ/OneDrive - Aalborg Universitet/IBD project/V1/data/"

```

# Transform MetaPhlAn output into a phloseq object

Function for transforming MetaPhlAn output into a phyloseq object:
```{r}
metaphlanToPhyloseq <- function(
  tax,
  metadat=NULL,
  simplenames=TRUE,
  roundtointeger=FALSE,
  split="|"){
  ## tax is a matrix or data.frame with the table of taxonomic abundances, rows are taxa, columns are samples
  ## metadat is an optional data.frame of specimen metadata, rows are samples, columns are variables
  ## if simplenames=TRUE, use only the most detailed level of taxa names in the final object
  ## if roundtointeger=TRUE, values will be rounded to the nearest integer
  xnames = rownames(tax)
  shortnames = gsub(paste0(".+\\", split), "", xnames)
  if(simplenames){
    rownames(tax) = shortnames
  }
  if(roundtointeger){
    tax = round(tax * 1e4)
  }
  x2 = strsplit(xnames, split=split, fixed=TRUE)
  taxmat = matrix(NA, ncol=max(sapply(x2, length)), nrow=length(x2))
  colnames(taxmat) = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species", "Strain")[1:ncol(taxmat)]
  rownames(taxmat) = rownames(tax)
  for (i in 1:nrow(taxmat)){
    taxmat[i, 1:length(x2[[i]])] <- x2[[i]]
  }
  taxmat = gsub("[a-z]__", "", taxmat)
  taxmat = phyloseq::tax_table(taxmat)
  otutab = phyloseq::otu_table(tax, taxa_are_rows=TRUE)
  if(is.null(metadat)){
    res = phyloseq::phyloseq(taxmat, otutab)
  }else{
    res = phyloseq::phyloseq(taxmat, otutab, phyloseq::sample_data(metadat))
  }
  return(res)
}
```

Read in the metadata: 
```{r echo=T, message=F, error=F, warning=F}
meta <-  read_delim(file = paste0(path, "metadata/Tilpasset_metadata.txt"), col_names = T, delim= "\t")
meta <- as_tibble(meta)
dim(meta)

#Make column with categorical clinical disease score for both children and adults
meta <- meta %>% mutate(score_p = ifelse(pucai_current < 10, "remission", ifelse(pucai_current <= 34, "mild", ifelse(pucai_current<=64, "moderate", ifelse(pucai_current>64, "severe", "NA")))))

meta <- meta %>% mutate(score_s = ifelse(sccai_current<3, "remission", ifelse(sccai_current<=5, "mild", ifelse(sccai_current<=11, "moderate", ifelse(sccai_current>11, "severe", "NA")))))

meta <- meta %>% mutate(score = ifelse(is.na(score_p), score_s, ifelse(is.na(score_s), score_p, "NA")))

#Make discrete numeric disease score:
meta$score_num <- ifelse(meta$score =="remission", 1, ifelse(meta$score=="mild", 2, ifelse(meta$score=="moderate",3, 4)))

#Make binary disease scores:
meta$score <- factor(meta$score, levels = c("remission", "mild", "moderate", "severe"))
meta$score_binary <- as.factor(ifelse(meta$score == "remission", 0, 1))

meta$fcal_binary <- as.factor(ifelse(meta$f_cal_current < 250, 0, 1))

#Read in the total read depth and add to meta data:
qc_stats <-  read_delim(file = paste0(path, "../analysis/summary_stats_of_QC.txt"), col_names = T, delim= "\t")
qc_stats$rdepth <- as.numeric(gsub(",", "", qc_stats$clean_R1))*2
qc_stats <- qc_stats %>% dplyr::select(J_ID, rdepth)
meta <- merge(meta, qc_stats, by.x="J_ID", by.y="J_ID")

#Save meta data:
write.table(meta, paste0(path, "metadata/", "preped_metadata.txt"), row.names = F, sep="\t")

#Prepare for phyloseq
meta <- as.data.frame(meta)
rownames(meta) = meta$J_ID
```

Investigate summary statistics of meta data:
```{r}
meta_total <- read.table(paste0(path, "metadata/", "metadata_excel_new.txt"), sep="\t", header = T)
meta_total$BMI <-  meta_total$weight/(meta_total$height/100)^2

#Total number of patients:
dim(meta_total)

#Total number of children and adults
adults <- meta_total %>% filter(age_gr == 2)
children <- meta_total %>% filter(age_gr == 1)
dim(adults)
dim(children)

#Split by gender:
dim(children %>% filter(sex==0))
dim(children %>% filter(sex==1))
dim(adults %>% filter(sex==0))
dim(adults %>% filter(sex==1))

#Study group:
dim(children %>% filter(study_gr==1))
dim(children %>% filter(study_gr==2))
dim(adults %>% filter(study_gr==1))
dim(adults %>% filter(study_gr==2))

#Treatment before baseline:
dim(children %>% filter(asa_0 == 1))
dim(adults %>% filter(asa_0 == 1))

dim(children %>% filter(pred_0 == 1))
dim(adults %>% filter(pred_0 == 1))

dim(children %>% filter(aza_0 == 1))
dim(adults %>% filter(aza_0 == 1))

dim(children %>% filter(bio_0 == 1))
dim(adults %>% filter(bio_0 == 1))

dim(children %>% filter(l_asa_0 == 1))
dim(adults %>% filter(l_asa_0 == 1))

dim(children %>% filter(l_pred_0 == 1))
dim(adults %>% filter(l_pred_0 == 1))

dim(children %>% filter(asa_0 ==0 &pred_0 == 0 &aza_0 == 0 &bio_0 == 0 &l_asa_0 == 0 & l_pred_0 == 0))
dim(adults %>% filter(asa_0 ==0 &pred_0 == 0 &aza_0 == 0 &bio_0 == 0 &l_asa_0 == 0 & l_pred_0 == 0))

```

Read in the MetaPhlan count data and transform it into phyloseq object:
```{r}
#Count data
bug_count <- read.csv(paste0(path, "humann3_processed_output/Joined_tables/", "Joined_metaphlan_bugs_list_counts.tsv"), sep = "\t", strip.white = T, stringsAsFactors = F, row.names = 1)

#Change J-ID
colnames(bug_count) <- substr(colnames(bug_count), start = 1, stop = 6)

#Subtract species level to include in phyloseq
level.letter = 's'
species_count = as.data.frame(matrix(nrow=0, ncol=ncol(bug_count)))

colnames(species_count) <- colnames(bug_count)
  
mph.names = rownames(bug_count)
    
for (i in 1:length(mph.names)){
    A = unlist(strsplit(mph.names[i], "|", fixed=TRUE))
    mph.genera.full = A[length(A)]
    B = unlist(strsplit(A[length(A)], "_", fixed=TRUE))
    if (B[1] == level.letter){
      #print(mph.names[i])
      species_count[mph.names[i],] = bug_count[mph.names[i],]
      }
    }

#Change setup of table
ps_count <- metaphlanToPhyloseq(species_count, metadat=meta)
otu_table(ps_count)[1:5,10:14]
tax_table(ps_count)[1:2,]
ps_count #208 samples, 550 taxa

#Save phyloseq object
saveRDS(ps_count, paste0(path, "humann3_processed_output/Phyloseq/","ps_count.rds"))

```

Read in relative abundance MetaPhlan data and transform it into phyloseq object, like what was done with the count data:
```{r}
#Relative abundance data
bug_relab <- read.csv(paste0(path, "humann3_processed_output/Joined_tables/", "Joined_metaphlan_bugs_list_relab.tsv"), sep = "\t", strip.white = T, stringsAsFactors = F, row.names = 1) 

#Change J-ID and save UNKNOWN abundance
colnames(bug_relab) <- substr(colnames(bug_relab), start = 1, stop = 6)
relab_unknown <- bug_relab[1,]

#Subtract species level to include in phyloseq
level.letter = 's'
species_relab = as.data.frame(matrix(nrow=0, ncol=ncol(bug_relab)))
colnames(species_relab) <- colnames(bug_relab)
    
mph.names = rownames(bug_relab)
for (i in 1:length(mph.names)){
    A = unlist(strsplit(mph.names[i], "|", fixed=TRUE))
    mph.genera.full = A[length(A)]
    B = unlist(strsplit(A[length(A)], "_", fixed=TRUE))
    if (B[1] == level.letter){
      #print(mph.names[i])
      species_relab[mph.names[i],] = bug_relab[mph.names[i],]
    }
  }
dim(species_relab)

#This does not summarise to 100, because unmapped are subtracted from the data frame:
colSums(species_relab)[1:10]

#Change setup of table
ps_relab <- metaphlanToPhyloseq(species_relab, metadat = meta)
otu_table(ps_relab)[1:5,10:14]
tax_table(ps_relab)[1:2,]
ps_relab #208 samples, 550 taxa

#Save phyloseq object
saveRDS(ps_relab, paste0(path, "humann3_processed_output/Phyloseq/","ps_relab.rds"))
```

# Quality control based on read depth

Remove samples, if the have a read depth below 10,000 reads:
```{r}
#Histogram of sequencing depth
ggplot()+
  geom_histogram(mapping=aes(sample_sums(ps_count)), fill = "lightskyblue1", color="black", bins = 100)+
  theme_classic()+xlab("Read depth")+ylab("Count")+
  geom_vline(xintercept = 10000, linetype = "dashed")
    
min(sample_sums(ps_count))

#Remove sample with read depth below 10,000:
ps_sub = prune_samples(sample_sums(ps_count)>=10000, ps_count)
ps_sub #207 samples - one was removed

saveRDS(ps_sub, paste0(path, "humann3_processed_output/Phyloseq/","ps_count_filtered.rds"))

#Which sample was removed? It should also be removed in the data set with relative abundance:
removed <- setdiff(sample_names(ps_count), sample_names((ps_sub)))
removed #J28550
ps_relab_sub <- subset_samples(ps_relab, !(J_ID %in% removed))
ps_relab_sub #207 samples and 550 taxa
    
saveRDS(ps_relab_sub,paste0(path, "humann3_processed_output/Phyloseq/","ps_relab_filtered.rds"))
```



# Inspection of data using stacked barplots

Make stacked barplots to inspect the data. Normally, the abundance in the plots will summarise to 100. But in this setup, I have not excluded the propotion of unmapped reads. Therefore, the stacked bar plots look different than usual.
```{r}
ps_relab <- readRDS(paste0(path, "humann3_processed_output/Phyloseq/", "ps_relab_filtered.rds"))

#Based on Kingdom (bacteria/eukaryote/archae)
temp_1 <- ps_relab %>% tax_glom(taxrank = "Kingdom") %>% psmelt()

ggplot(temp_1)+
  geom_bar(aes(x=Sample, y= Abundance, fill = Kingdom), stat = "identity", position = "stack")+
  xlab("Participants")+theme_classic()+ theme(axis.text.x=element_blank())

#Based on Phylum
tax_level <- "Phylum"
temp <- ps_relab %>% tax_glom(taxrank = tax_level) %>% psmelt() %>% filter(Abundance > 2)

ggplot(temp)+
  geom_bar(aes(x=Sample, y= Abundance, fill = Phylum), stat = "identity", position = "stack")+
  xlab("Participants")+theme_classic()+ theme(axis.text.x=element_blank())


#Based on Class:
temp <- ps_relab %>% tax_glom(taxrank = "Class") %>% psmelt() %>% filter(Abundance > 2)

#Divided into sex:
ggplot(temp)+
  geom_bar(aes(x=Sample, y= Abundance, fill = Class), stat = "identity", position = "stack")+
  xlab("Participants")+theme_classic()+ theme(axis.text.x=element_blank())+facet_wrap(~sex, scales="free_x")

#Divided by age group and study group:
temp$age_gr <- ifelse(temp$age_gr == 1, "Pediatric", "Adult")
temp$study_gr <- ifelse(temp$study_gr == 1, "Newly diagnosed", "Older diagnosis")
ggplot(temp)+
  geom_bar(aes(x=Sample, y= Abundance, fill = Class), stat = "identity", position = "stack")+
  xlab("Participants")+theme_classic()+ theme(axis.text.x=element_blank())+facet_wrap(~age_gr+study_gr, scales="free_x")

#Based on Order:
temp <- ps_relab %>% tax_glom(taxrank = "Order") %>% psmelt() %>% filter(Abundance > 2)
#Divided into sex:
ggplot(temp)+
  geom_bar(aes(x=Sample, y= Abundance, fill = Order), stat = "identity", position = "stack")+
  xlab("Participants")+theme_classic()+ theme(axis.text.x=element_blank())+facet_wrap(~sex, scales="free_x")
#Divided by age group:
ggplot(temp)+
  geom_bar(aes(x=Sample, y= Abundance, fill = Order), stat = "identity", position = "stack")+
  xlab("Participants")+theme_classic()+ theme(axis.text.x=element_blank())+facet_wrap(~age_gr, scales="free_x")

#Based on Family:
temp <- ps_relab %>% tax_glom(taxrank = "Family") %>% psmelt() %>% filter(Abundance > 2)
#Divided into sex:
ggplot(temp)+
  geom_bar(aes(x=Sample, y= Abundance, fill = Family), stat = "identity", position = "stack")+
  xlab("Participants")+theme_classic()+ theme(axis.text.x=element_blank())+facet_wrap(~sex, scales="free_x")
#Divided by age group:
ggplot(temp)+
  geom_bar(aes(x=Sample, y= Abundance, fill = Family), stat = "identity", position = "stack")+
  xlab("Participants")+theme_classic()+ theme(axis.text.x=element_blank())+facet_wrap(~age_gr, scales="free_x")
```

For Genus and Species plots, the legend is not shown, is it would be too large:
```{r}
#Based on Genus:
temp <- ps_relab %>% tax_glom(taxrank = "Genus") %>% psmelt() %>% filter(Abundance > 2)
#Divided into sex:
ggplot(temp)+
  geom_bar(aes(x=Sample, y= Abundance, fill = Genus), stat = "identity", position = "stack")+
  xlab("Participants")+theme_classic()+ theme(axis.text.x=element_blank())+facet_wrap(~sex, scales="free_x")+theme(legend.position = "none")
#Divided by age group:
ggplot(temp)+
  geom_bar(aes(x=Sample, y= Abundance, fill = Genus), stat = "identity", position = "stack")+
  xlab("Participants")+theme_classic()+ theme(axis.text.x=element_blank())+facet_wrap(~age_gr, scales="free_x")+theme(legend.position = "none")

#Based on Species:
temp <- ps_relab %>% tax_glom(taxrank = "Species") %>% psmelt() %>% filter(Abundance > 2)
#Divided into sex:
ggplot(temp)+
  geom_bar(aes(x=Sample, y= Abundance, fill = Species), stat = "identity", position = "stack")+
  xlab("Participants")+theme_classic()+ theme(axis.text.x=element_blank())+facet_wrap(~sex, scales="free_x")+theme(legend.position = "none")
#Divided by age group:
ggplot(temp)+
  geom_bar(aes(x=Sample, y= Abundance, fill = Species), stat = "identity", position = "stack")+
  xlab("Participants")+theme_classic()+ theme(axis.text.x=element_blank())+facet_wrap(~age_gr, scales="free_x")+theme(legend.position = "none")

```

# Calculation of alpha diversity

Calculate alpha diversity using phyloseq package:
```{r warning=FALSE}
ps_count <- readRDS(paste0(path, "humann3_processed_output/Phyloseq/","ps_count_filtered.rds"))

#Calculate alpha div based on counts
alpha_div <-estimate_richness(ps_count, measures=c("Chao1", "Shannon", "InvSimpson", "Observed"))
head(alpha_div)

#Create matrix
alpha_div$J_ID <- row.names(alpha_div)

# Save alpha diversity
write.table(alpha_div, paste0(path, "humann3_processed_output/processed_tables/",'AlphaDiv.txt'), col.names = T, row.names = F, quote = F, sep = '\t')

```

Calculate alpha diversity based on relative abundance and investigate the correlation:
```{r}
matrix_spec <- t(otu_table(ps_relab))
div_shannon_003 <- diversity(matrix_spec, index="shannon")  ### Shannon entropy
spec_nr_003 <- specnumber(matrix_spec,MARGIN=1)             ### Species number (observed)

alpha_div_relab <- as.data.frame(cbind(spec_nr_003,div_shannon_003))
alpha_div_relab$J_ID <- row.names(alpha_div_relab)

# Save alpha diversity
write.table(alpha_div_relab, paste0(path, "humann3_processed_output/processed_tables/",'AlphaDiv_relab.txt'), col.names = T, row.names = F, quote = F, sep = '\t')


# Compare the two methods
alpha_total <- merge(alpha_div, alpha_div_relab, by.x="J_ID", by.y="J_ID")
head(alpha_total)

alpha_cor <- alpha_total %>% dplyr::select(-J_ID)
#Spearman correlation
df1 <- cor(alpha_cor, method = "spearman", use = "pairwise.complete.obs")
df1
#Pearson correlation:
df2 <- cor(alpha_cor, method = "pearson", use = "pairwise.complete.obs")
df2
```


# Calculation of beta diversity 

Calculate beta diversity based on species level:
```{r warning=FALSE}
ps_count <- readRDS(paste0(path, "humann3_processed_output/Phyloseq/","ps_count_filtered.rds"))
ps_relab <- readRDS(paste0(path, "humann3_processed_output/Phyloseq/","ps_relab_filtered.rds"))

#Bray Curtis: 
beta_div_relab_bray = phyloseq::distance(ps_relab, method='bray')
beta_div_relab_bray %>% as.vector %>% summary

#Jaccard:
beta_div_relab_jaccard = phyloseq::distance(ps_relab, method='jaccard', binary = TRUE)
beta_div_relab_jaccard %>% as.vector %>% summary


#Save files
saveRDS(beta_div_relab_bray, paste0(path,"humann3_processed_output/processed_tables/", "Beta.diversity_relab_bray.RDS"))
saveRDS(beta_div_relab_jaccard, paste0(path,"humann3_processed_output/processed_tables/", "Beta.diversity_relab_jaccard.RDS"))

```


Plot beta diversity using NMDS to check for outliers:
Can also be used to visually inspect association with variables. 
Stress:
<0.05 = excellent
<0.1 = great
<0.2 = good/ok
<0.3 = poor

Bray Curtis distance on species level:
```{r}
BRAY_NMDS_relab_SPE=metaMDS(beta_div_relab_bray, k=2,trymax=30)
BRAY_NMDS_relab_SPE$stress

#Plotting
data.scores = as.data.frame(scores(BRAY_NMDS_relab_SPE))
data.scores$sex <- sample_data(ps_relab)$sex
data.scores$age_gr <- sample_data(ps_relab)$age_gr
data.scores$time_point <- sample_data(ps_relab)$time_point
data.scores$f_cal <- sample_data(ps_relab)$f_cal_current
data.scores$f_cal_binary <- sample_data(ps_relab)$fcal_binary
data.scores$score <- sample_data(ps_relab)$score
data.scores$score_binary <- sample_data(ps_relab)$score_binary

ggplot(data=data.scores, aes(x=NMDS1, y=NMDS2, color = as.factor(sex)))+
  geom_point()+theme_classic()+
  stat_ellipse()+labs(color="Sex")

ggplot(data=data.scores, aes(x=NMDS1, y=NMDS2, color = as.factor(age_gr)))+
  geom_point()+theme_classic()+
  stat_ellipse()+labs(color="Age group")

data.scores1 <- data.scores %>% filter(!is.na(f_cal))
ggplot(data=data.scores1, aes(x=NMDS1, y=NMDS2, color = f_cal))+
  geom_point(size=1)+theme_classic()+
  labs(color="Faecal calprotectin")

data.scores1$f_cal_binary1 <- factor(ifelse(data.scores1$f_cal_binary==0, "Remission", "Disease"), levels=c("Remission", "Disease"))
ggplot(data=data.scores1, aes(x=NMDS1, y=NMDS2, color = f_cal_binary1))+
  geom_point(size=1)+theme_classic()+
  labs(color="Paraclinical")+
  stat_ellipse()+ggtitle("Bray Curtis distance")

data.scores2 <- data.scores %>% filter(!is.na(score))
ggplot(data=data.scores2, aes(x=NMDS1, y=NMDS2, color = score))+
  geom_point(size=1)+theme_classic()+
  labs(color="Disease score")+
  stat_ellipse()

data.scores2$score_binary1 <- factor(ifelse(data.scores2$score_binary==0, "Remission", "Disease"), levels=c("Remission", "Disease"))
ggplot(data=data.scores2, aes(x=NMDS1, y=NMDS2, color = score_binary1))+
  geom_point(size=1)+theme_classic()+
  labs(color="Clinical")+
  stat_ellipse()+ggtitle("Bray Curtis distance")

```

Jaccard distance on species level:
```{r}

JAC_NMDS_relab_SPE=metaMDS(beta_div_relab_jaccard,k=2,trymax=30)
JAC_NMDS_relab_SPE$stress 
    
#Plotting
data.scores = as.data.frame(scores(JAC_NMDS_relab_SPE))
data.scores$sex <- sample_data(ps_relab)$sex
data.scores$time_point <- sample_data(ps_relab)$time_point
data.scores$age_gr <- sample_data(ps_relab)$age_gr
data.scores$f_cal <- sample_data(ps_relab)$f_cal_current
data.scores$f_cal_binary <- sample_data(ps_relab)$fcal_binary
data.scores$score <- sample_data(ps_relab)$score
data.scores$score_binary <- sample_data(ps_relab)$score_binary

ggplot(data=data.scores, aes(x=NMDS1, y=NMDS2, color = as.factor(sex)))+
  geom_point()+theme_classic()+
  stat_ellipse()+labs(color="Sex")

ggplot(data=data.scores, aes(x=NMDS1, y=NMDS2, color = as.factor(age_gr)))+
  geom_point()+theme_classic()+
  stat_ellipse()+labs(color="Age group")

data.scores1 <- data.scores %>% filter(!is.na(f_cal))
ggplot(data=data.scores1, aes(x=NMDS1, y=NMDS2, color = f_cal))+
  geom_point(size=1)+theme_classic()+
  labs(color="Faecal calprotectin")

data.scores1$f_cal_binary1 <- factor(ifelse(data.scores1$f_cal_binary==0, "Remission", "Disease"), levels=c("Remission", "Disease"))
ggplot(data=data.scores1, aes(x=NMDS1, y=NMDS2, color = f_cal_binary1))+
  geom_point(size=1)+theme_classic()+
  labs(color="Paraclinical")+
  stat_ellipse()+ggtitle("Jaccard distance")

data.scores2 <- data.scores %>% filter(!is.na(score))
ggplot(data=data.scores2, aes(x=NMDS1, y=NMDS2, color = score))+
  geom_point(size=1)+theme_classic()+
  labs(color="Disease score")+
  stat_ellipse()

data.scores2$score_binary1 <- factor(ifelse(data.scores2$score_binary==0, "Remission", "Disease"), levels=c("Remission", "Disease"))
ggplot(data=data.scores2, aes(x=NMDS1, y=NMDS2, color = score_binary1))+
  geom_point(size=1)+theme_classic()+
  labs(color="Clinical")+
  stat_ellipse()+ggtitle("Jaccard distance")


```


# Prefiltering of dataframe for single species analyses

Select species to test in single species analyses. Species are preselected based on high mean and high variance across samples. This is done to increase power and only investigate species, that could be biological interesting. Finally, species are also selected based on proportion of zeros:
```{r}
ps_count <- readRDS(paste0(path, "humann3_processed_output/Phyloseq/","ps_count_filtered.rds"))
ps_relab <- readRDS(paste0(path, "humann3_processed_output/Phyloseq/","ps_relab_filtered.rds"))

count <- as.data.frame(t(otu_table(ps_count)))
relab <- as.data.frame(t(otu_table(ps_relab)))

#Relab data
#Subset based on highest mean and variance
step1 <- relab[ ,colMeans(relab) >= quantile(colMeans(relab), 0.5)]
dim(step1) #reduced to 275 taxa
                
step2 <- apply(step1, 2, var)
step2 <- step1[ ,step2 >= quantile( step2, 0.50) ]
dim(step2) #reduced to 138 taxa

ps_relab_sub_mean_var <- prune_taxa(colnames(step2),ps_relab)
saveRDS(ps_relab_sub_mean_var, paste0(path, "humann3_processed_output/Phyloseq/", "ps_relab_50mean_50var.rds"))

#Subset based on zero proportion:
non <- nonzero(step2)
for_analysis <- non$nonzero.p[non$nonzero.p>0.25]
step3 <- step2[,names(for_analysis)]
dim(step3) #reduced to 75 taxa

ps_relab_sub_mean_var_zero <- prune_taxa(colnames(step3),ps_relab)
saveRDS(ps_relab_sub_mean_var_zero, paste0(path, "humann3_processed_output/Phyloseq/", "ps_relab_50mean_50var_zero.rds"))

##Keep the same taxa in the count dataset
ps_count_sub_mean_var <- prune_taxa(colnames(step2),ps_count)
saveRDS(ps_count_sub_mean_var, paste0(path, "humann3_processed_output/Phyloseq/", "ps_count_50mean_50var.rds"))

ps_count_sub_mean_var_zero <- prune_taxa(colnames(step3),ps_count)
saveRDS(ps_count_sub_mean_var_zero, paste0(path, "humann3_processed_output/Phyloseq/", "ps_count_50mean_50var_zero.rds"))

#Evaluate: Zeros, dispersion
#Mean:
mean(apply(otu_table(ps_count_sub_mean_var_zero), 1, FUN=mean))
#Variance:
mean(apply(otu_table(ps_count_sub_mean_var_zero), 1, FUN=stats::var))
#Zero proportion:
mean(apply(otu_table(ps_count_sub_mean_var_zero), 1, function(x) sum(x==0)/207))
sd(apply(otu_table(ps_count_sub_mean_var_zero), 1, function(x) sum(x==0)/207))

#Mean to variance ratio:
mean(apply(otu_table(ps_count_sub_mean_var_zero), 1, function(x) var(x)/mean(x)))
sd(apply(otu_table(ps_count_sub_mean_var_zero), 1, function(x) var(x)/mean(x)))

```


# Plots for manuscript: Figure 1 and supplementary figure 1.
```{r}

#Figure 1C, Fecal, stacked barplot:

#Renormalize to 100%:
ps_relab <- readRDS(paste0(path, "humann3_processed_output/Phyloseq/","ps_relab_filtered2.rds"))

ps_relab2 <- transform_sample_counts(ps_relab, function(x) x / sum(x) ) #100% in decimals

temp <- ps_relab2 %>% tax_glom(taxrank = "Class") %>% psmelt() %>% filter(Abundance > 0.02)

temp$age_gr <- ifelse(temp$age_gr == 1, "Pediatric", "Adult")
temp$study_gr <- ifelse(temp$study_gr == 1, "Newly diagnosed", "Older diagnosis")
fig_1c <- ggplot(temp)+
  geom_bar(aes(x=Sample, y= Abundance, fill = Class), stat = "identity", position = "stack")+
  xlab("Participants")+theme_classic()+ theme(axis.text.x=element_blank())+facet_wrap(~age_gr+study_gr, scales="free_x")+
  theme(text = element_text(family = "sans", size = 8))

ggsave(fig_1c, filename = "C:/Users/FX76TZ/OneDrive - Aalborg Universitet/IBD project/V1/illustrations/Manuscript/Figure_1C.tiff", width = 130, height = 110, units = "mm", dpi=300)



# Figure 1D:
set.seed(0)
beta_div_relab_bray = phyloseq::distance(ps_relab, method='bray')
BRAY_NMDS_relab_SPE=metaMDS(beta_div_relab_bray, k=2,trymax=30)
BRAY_NMDS_relab_SPE$stress

data.scores = as.data.frame(scores(BRAY_NMDS_relab_SPE))
data.scores$study_gr <- sample_data(ps_relab)$study_gr
data.scores$age_gr <- sample_data(ps_relab)$age_gr
data.scores$time_point <- sample_data(ps_relab)$time_point

data.scores$Group <- ifelse(data.scores$age_gr == 1 & data.scores$study_gr == 1, "Pediatric: Newly diagnosed", ifelse(data.scores$age_gr == 1 & data.scores$study_gr == 2,"Pediatric: Older diagnosis", ifelse(data.scores$study_gr == 1, "Adult: Newly diagnosed", "Adult: Older diagnosis")))

fig_1d <- ggplot(data=data.scores, aes(x=NMDS1, y=NMDS2, color = Group))+
  geom_point(size=1)+theme_classic()+
  stat_ellipse()+
  theme(text = element_text(family = "sans", size = 8))

ggsave(fig_1d, filename = "C:/Users/FX76TZ/OneDrive - Aalborg Universitet/IBD project/V1/illustrations/Manuscript/Figure_1D.tiff", width = 110, height = 60, units = "mm", dpi=300)


# Figure 1A: Biopsies, Stacked barplots
main.dir = "C:/Users/FX76TZ/OneDrive - Aalborg Universitet/IBD project/UC - biopsier/"
setwd(paste0( file.path( main.dir, "data/16S/data_processed")))
ps_relab = readRDS("DATA.ps.raw_SampleFilter.relab_meta.modified.rds")
 
temp <- ps_relab %>% tax_glom(taxrank = "Class") %>% psmelt() %>% filter(Abundance > 0.02)

temp$study_gr <- ifelse(temp$study_gr == 1, "Newly diagnosed", "Older diagnosis")
temp$GI.location <- ifelse(temp$GI.location == "c", "Caecum", ifelse(temp$GI.location == "r", "Rectum", ifelse(temp$GI.location == "s", "Sigmoid colon", "Tansverse colon")))

fig_supp <- ggplot(temp)+
  geom_bar(aes(x=Sample, y= Abundance, fill = Class), stat = "identity", position = "stack")+
  xlab("Participants")+theme_classic()+ theme(axis.text.x=element_blank())+facet_wrap(~GI.location, scales="free_x")+
  theme(text = element_text(family = "sans", size = 8))

fig_1a <- ggplot(temp)+
  geom_bar(aes(x=Sample, y= Abundance, fill = Class), stat = "identity", position = "stack")+
  xlab("Participants")+theme_classic()+ theme(axis.text.x=element_blank())+facet_wrap(~age_gr+study_gr, scales="free_x")+
  theme(text = element_text(family = "sans", size = 8))

ggsave(fig_1a, filename = "C:/Users/FX76TZ/OneDrive - Aalborg Universitet/IBD project/V1/illustrations/Manuscript/Figure_1A.tiff", width = 130, height = 110, units = "mm", dpi=300)

ggsave(fig_supp, filename = "C:/Users/FX76TZ/OneDrive - Aalborg Universitet/IBD project/V1/illustrations/Manuscript/Figure_Supplement.tiff", width = 130, height = 110, units = "mm", dpi=300)


#Figure 1B:
set.seed(0)
beta_div_relab_bray = phyloseq::distance(ps_relab, method='bray')
BRAY_NMDS_relab_SPE=metaMDS(beta_div_relab_bray, k=2,trymax=30)
BRAY_NMDS_relab_SPE$stress

data.scores = as.data.frame(scores(BRAY_NMDS_relab_SPE))
data.scores$study_gr <- sample_data(ps_relab)$study_gr
data.scores$age_gr <- sample_data(ps_relab)$age_gr
data.scores$time_point <- sample_data(ps_relab)$time_point
data.scores$GI.location <- sample_data(ps_relab)$GI.location

data.scores$Group <- ifelse(data.scores$age_gr == "Paediatric" & data.scores$study_gr == 1, "Pediatric: Newly diagnosed", ifelse(data.scores$age_gr == "Paediatric" & data.scores$study_gr == 2,"Pediatric: Older diagnosis", ifelse(data.scores$study_gr == 1, "Adult: Newly diagnosed", "Adult: Older diagnosis")))

fig_1b <- ggplot(data=data.scores, aes(x=NMDS1, y=NMDS2, color = Group))+
  geom_point(size=1)+theme_classic()+
  stat_ellipse()+
  theme(text = element_text(family = "sans", size = 8))

ggsave(fig_1b, filename = "C:/Users/FX76TZ/OneDrive - Aalborg Universitet/IBD project/V1/illustrations/Manuscript/Figure_1B.tiff", width = 110, height = 60, units = "mm", dpi=300)



#Gather figure 1:
p_left <- ggarrange(fig_1a, fig_1c, nrow=2, labels = c("A)", "C)"), font.label = list(size=10, family="sans"), heights = c(1, 1))

p_right <- ggarrange(fig_1b, fig_1d, nrow=2, labels=c("B)", "D)"), common.legend = T, legend = "right", font.label = list(size=10, family="sans"), heights =  c(1, 1))

p_total <- ggarrange(p_left, p_right, ncol=2, widths =  c(1,0.8))

ggsave(p_total, filename = "C:/Users/FX76TZ/OneDrive - Aalborg Universitet/IBD project/V1/illustrations/Manuscript/Figure_1.tiff", width = 174, height = 200, units = "mm", dpi=300)



#Figure supp 1B
set.seed(0)
beta_div_relab_bray = phyloseq::distance(ps_relab, method='bray')
BRAY_NMDS_relab_SPE=metaMDS(beta_div_relab_bray, k=2,trymax=30)
BRAY_NMDS_relab_SPE$stress

data.scores = as.data.frame(scores(BRAY_NMDS_relab_SPE))
data.scores$GI.location <- sample_data(ps_relab)$GI.location
data.scores$GI.location <- ifelse(data.scores$GI.location == "c", "Caecum", ifelse(data.scores$GI.location == "r", "Rectum", ifelse(data.scores$GI.location == "s", "Sigmoid colon", "Tansverse colon")))

fig_sup1b <- ggplot(data=data.scores, aes(x=NMDS1, y=NMDS2, color = GI.location))+
  geom_point(size=1)+theme_classic()+
  stat_ellipse()+guides(color=guide_legend(title="GI location"))+
  theme(text = element_text(family = "sans", size = 8))

#Figure supp 1C
set.seed(0)
jac_div_relab_bray = phyloseq::distance(ps_relab, method='jaccard', binary=T)
JAC_NMDS_relab_SPE=metaMDS(jac_div_relab_bray, k=2,trymax=30)
JAC_NMDS_relab_SPE$stress

data.scores = as.data.frame(scores(JAC_NMDS_relab_SPE))
data.scores$GI.location <- sample_data(ps_relab)$GI.location
data.scores$GI.location <- ifelse(data.scores$GI.location == "c", "Caecum", ifelse(data.scores$GI.location == "r", "Rectum", ifelse(data.scores$GI.location == "s", "Sigmoid colon", "Tansverse colon")))

fig_sup1c <- ggplot(data=data.scores, aes(x=NMDS1, y=NMDS2, color = GI.location))+
  geom_point(size=1)+theme_classic()+
  stat_ellipse()+guides(color=guide_legend(title="GI location"))+
  theme(text = element_text(family = "sans", size = 8))

#Figure supp 1A
matrix_spec <- t(otu_table(ps_relab))
div_shannon_003 <- diversity(t(matrix_spec), index="shannon")  ### Shannon entropy
spec_nr_003 <- specnumber(matrix_spec,MARGIN=2)             ### Species number (observed)
meta <- as.data.frame(as.matrix(sample_data(ps_relab)))
meta$J_ID <- row.names(meta)

alpha_div_relab <- as.data.frame(cbind(spec_nr_003,div_shannon_003))
alpha_div_relab$J_ID <- row.names(alpha_div_relab)
alpha_div <- merge(alpha_div_relab, meta, by.x="J_ID", by.y="J_ID")
alpha_div$GI.location <- ordered(alpha_div$GI.location, levels = c('r','s',  't', 'c'), labels = c('r','s',  't', 'c'))
head(alpha_div_relab)
alpha_div$GI.location <- ifelse(alpha_div$GI.location == "c", "Caecum", ifelse(alpha_div$GI.location == "r", "Rectum", ifelse(alpha_div$GI.location == "s", "Sigmoid colon", "Tansverse colon")))

fig_sup1a <- ggplot(data=alpha_div)+
  geom_boxplot(aes(x=GI.location, y=div_shannon_003))+xlab("")+ylab("Shannon diversity")+theme_classic()


#Gather figure supp 1:
p_bottom <- ggarrange(fig_sup1b, fig_sup1c, common.legend =T, legend = "bottom", labels=c("B)","C)"), font.label = list(size=10, family="sans"))

p_total <- ggarrange(fig_sup1a, p_bottom, nrow=2, heights = c(1,1))

ggsave(p_total, filename = "C:/Users/FX76TZ/OneDrive - Aalborg Universitet/IBD project/V1/illustrations/Manuscript/Figure_Sup1.tiff", width = 174, height = 145, units = "mm", dpi=300)

```