---
title: "Investigation of diversity and stability"
output:
  github_document:
    toc: true
editor_options:
  chunk_output_type: console
---
# Read in packages and data

Read in packages and path to folder:
```{r echo=T, message=F, error=F, warning=F}
library(dplyr)
library(ggplot2)
library(readr)
library(ggpubr)
library(vegan)
library(caret)
library(phyloseq)
library(lme4)
library(lmerTest)
library(nlme)
library(RCurl)
library(MuMIn)
path = "C:/Users/FX76TZ/OneDrive - Aalborg Universitet/IBD project/V1/data/"
```

Read in data:
```{r echo=T, message=F, error=F, warning=F}
#Meta data:
meta <- read.delim(paste0(path, "metadata/", "preped_metadata_filtered.txt"))

#Alpha diversity:
alpha_div <- read_delim(file=paste0(path, "humann3_processed_output/processed_tables/", "AlphaDiv_relab.txt"), delim="\t")
alpha_div <- merge(alpha_div, meta, by.x="J_ID", by.y="J_ID")
alpha_div$age_gr <- as.factor(alpha_div$age_gr)
alpha_div$study_gr <- as.factor(alpha_div$study_gr)

#Phyloseq objects:
ps_count <- readRDS(paste0(path, "humann3_processed_output/Phyloseq/","ps_count_filtered2.rds"))
ps_relab <- readRDS(paste0(path, "humann3_processed_output/Phyloseq/","ps_relab_filtered2.rds"))
sample_data(ps_count)$age_gr <- as.factor(sample_data(ps_count)$age_gr)
sample_data(ps_relab)$age_gr <- as.factor(sample_data(ps_relab)$age_gr)
sample_data(ps_count)$study_gr <- as.factor(sample_data(ps_count)$study_gr)
sample_data(ps_relab)$study_gr <- as.factor(sample_data(ps_relab)$study_gr)

```

Investigate if alpha diversity and unmapped reads are normally distributed and transform if not!:
```{r}
#Chao1
ggplot(data=alpha_div)+
  geom_histogram(aes(x=spec_nr_003), color="black", fill = "lightblue", bins=40)+theme_classic()+xlab("Observed diveristy")+ylab("Count")

#Shannon
ggplot(data=alpha_div)+
  geom_histogram(aes(x=div_shannon_003), color="black", fill = "lightblue", bins=40)+theme_classic()+xlab("Shannon diversity")+ylab("Count")

ggplot(data=alpha_div)+
  geom_histogram(aes(x=div_shannon_003^2), color="black", fill = "lightblue", bins=40)+theme_classic()+xlab("Squared Shannon diversity")+ylab("Count")

alpha_div$div_shannon_0032 <- (alpha_div$div_shannon_003)^2
```


# Tests of alpha diversity

Visual inspection of alpha diversity ~ disease score:

```{r}
#Clinical disease score:
alpha1 <- alpha_div %>% filter(!is.na(score))
alpha1$score <- factor(alpha1$score, levels=c("remission", "mild", "moderate", "severe"))

ggplot(alpha1)+
  geom_boxplot(aes(x=score, y=spec_nr_003, color=score))+theme_classic()+xlab("Disease score")+ylab("Observed diversity")+theme(legend.position = "none")

ggplot(alpha1)+
  geom_boxplot(aes(x=score, y=div_shannon_003, color=score))+theme_classic()+xlab("Disease score")+ylab("Shannon diversity")+theme(legend.position = "none")

#Paraclinical disease score
alpha2 <- alpha_div %>% filter(!is.na(f_cal_current))

ggplot(alpha2)+
  geom_point(aes(x=f_cal_current, y=spec_nr_003), size=1)+theme_classic()+xlab("Faecal calprotectin")+ylab("Observed diversity")+
  geom_vline(xintercept=250, color="blue", linetype = "dashed")

ggplot(alpha2)+
  geom_point(aes(x=f_cal_current, y=div_shannon_003), size=1)+theme_classic()+xlab("Faecal calprotectin")+ylab("Shannon diversity")+
  geom_vline(xintercept=250, color="blue", linetype = "dashed")
```

Tested using linear mixed models
```{r warning=FALSE}
divs <- c("spec_nr_003", "div_shannon_0032")

#Within study groups: Does alpha associate with age group?:
alpha_new <- alpha_div %>% filter(study_gr==1)
alpha_old <- alpha_div %>% filter(study_gr==2)

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr"))
  model1 <- lme(form, data=alpha_new, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ study_gr+age_gr"))
  model1 <- lme(form, data=alpha_div, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

#What is the association with disease score? Include age:
#Should interaction with age be included?
for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+f_cal_current*age_gr"))
  model1 <- lme(form, data=alpha_new, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+f_cal_current*age_gr"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}
for (div in divs){
  form <- as.formula(paste0(div, "~ study_gr+age_gr+f_cal_current*age_gr"))
  model1 <- lme(form, data=alpha_div, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

#Interaction is not significant. Fit without:

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+f_cal_current"))
  model1 <- lme(form, data=alpha_new, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+f_cal_current"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}
for (div in divs){
  form <- as.formula(paste0(div, "~ study_gr+age_gr+f_cal_current"))
  model1 <- lme(form, data=alpha_div, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}


#Split the dataset by age group and test influence of study group:
alpha_children <- alpha_div %>% filter(age_gr==1)
alpha_adults <- alpha_div %>% filter(age_gr==2)

for (div in divs){
  form <- as.formula(paste0(div, "~ study_gr"))
  model1 <- lme(form, data=alpha_children, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ study_gr"))
  model1 <- lme(form, data=alpha_adults, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+study_gr"))
  model1 <- lme(form, data=alpha_div, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

#Influence of medicine for earlier diagnosed:
for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+as.factor(bio)"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+as.factor(aza)"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+as.factor(pred)"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+as.factor(asa)"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

#Also include f-cal:
for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+f_cal_current+bio"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+f_cal_current+aza"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+f_cal_current+pred"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+f_cal_current+asa"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corAR1(form=~time_point|person_ID))
  print(summary(model1)$tTable)
}
```

# Tests of beta diversity
Tested with PERMANOVA
```{r}
#Number of samples/participant must be even! Select samples with 4 data points
ent <- subset_samples(ps_relab, !is.na(score))
df <- data.frame(sample_data(ent))
n_df <- df %>% group_by(person_ID) %>% summarise(n=n()) %>% filter(n>3)
df_reduced <- df %>% filter(person_ID %in% n_df$person_ID)
reduce <- df_reduced %>% group_by(person_ID)%>% arrange(time_point) %>% slice(1:4) %>% ungroup()
ps_reduced <- subset_samples(ps_relab, J_ID %in% reduce$J_ID)

#Divide by study group:
ps_new <- subset_samples(ps_reduced, study_gr==1)
ps_old <- subset_samples(ps_reduced, study_gr==2)

#Test age group- newly diagnosed:
perm <- how(within=Within(type="none"), plots=Plots(strata=sample_data(ps_new)$person_ID, type="free"), nperm=999)

set.seed(1)
adonis2(t(otu_table(ps_new))~age_gr, data=as.data.frame(as.matrix(sample_data(ps_new))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(t(otu_table(ps_new))~age_gr, data=as.data.frame(as.matrix(sample_data(ps_new))), method="jaccard", permutations=perm, by="margin", binary=T)

#Test age group- previously diagnosed:
perm <- how(within=Within(type="none"), plots=Plots(strata=sample_data(ps_old)$person_ID, type="free"), nperm=999)

set.seed(1)
adonis2(t(otu_table(ps_old))~age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(t(otu_table(ps_old))~age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="jaccard", permutations=perm, by="margin", binary=T)

#Test age group- full data set:
perm <- how(within=Within(type="none"), plots=Plots(strata=sample_data(ps_reduced)$person_ID, type="free"), nperm=999)

set.seed(1)
adonis2(t(otu_table(ps_reduced))~age_gr+study_gr, data=as.data.frame(as.matrix(sample_data(ps_reduced))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(t(otu_table(ps_reduced))~age_gr+study_gr, data=as.data.frame(as.matrix(sample_data(ps_reduced))), method="jaccard", permutations=perm, by="margin", binary=T)


#Test the influence of f-cal: Test first interaction with age:
ps_reduced <- subset_samples(ps_reduced, !is.na(f_cal_current))
ps_new <- subset_samples(ps_reduced, study_gr==1)
ps_old <- subset_samples(ps_reduced, study_gr==2)

#Newly diagnosed:
perm <- how(within=Within(type="free"), plots=Plots(strata=sample_data(ps_new)$person_ID, type="none"), nperm=999)

set.seed(1)
adonis2(t(otu_table(ps_new))~as.numeric(f_cal_current)+age_gr, data=as.data.frame(as.matrix(sample_data(ps_new))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(t(otu_table(ps_new))~as.numeric(f_cal_current)+age_gr, data=as.data.frame(as.matrix(sample_data(ps_new))), method="jaccard", permutations=perm, by="margin", binary=T)

#Previously diagnosed:
perm <- how(within=Within(type="free"), plots=Plots(strata=sample_data(ps_old)$person_ID, type="none"), nperm=999)

set.seed(1)
adonis2(t(otu_table(ps_old))~as.numeric(f_cal_current)+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(t(otu_table(ps_old))~as.numeric(f_cal_current)+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="jaccard", permutations=perm, by="margin", binary=T)

#Full dataset:
perm <- how(within=Within(type="free"), plots=Plots(strata=sample_data(ps_reduced)$person_ID, type="none"), nperm=999)

set.seed(1)
adonis2(t(otu_table(ps_reduced))~as.numeric(f_cal_current)+age_gr+study_gr, data=as.data.frame(as.matrix(sample_data(ps_reduced))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(t(otu_table(ps_reduced))~as.numeric(f_cal_current)+age_gr+study_gr, data=as.data.frame(as.matrix(sample_data(ps_reduced))), method="jaccard", permutations=perm, by="margin", binary=T)


# Divide by age group and test study group:
ps_children <- subset_samples(ps_reduced, age_gr==1)
ps_adults <- subset_samples(ps_reduced, age_gr==2)

#Children:
perm <- how(within=Within(type="none"), plots=Plots(strata=sample_data(ps_children)$person_ID, type="free"), nperm=999)

set.seed(1)
adonis2(t(otu_table(ps_children))~study_gr, data=as.data.frame(as.matrix(sample_data(ps_children))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(t(otu_table(ps_children))~study_gr, data=as.data.frame(as.matrix(sample_data(ps_children))), method="jaccard", permutations=perm, by="margin", binary=T)

#Adults:
perm <- how(within=Within(type="none"), plots=Plots(strata=sample_data(ps_adults)$person_ID, type="free"), nperm=999)

set.seed(1)
adonis2(t(otu_table(ps_adults))~study_gr, data=as.data.frame(as.matrix(sample_data(ps_adults))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(t(otu_table(ps_adults))~study_gr, data=as.data.frame(as.matrix(sample_data(ps_adults))), method="jaccard", permutations=perm, by="margin", binary=T)

#Full dataset:
perm <- how(within=Within(type="none"), plots=Plots(strata=sample_data(ps_reduced)$person_ID, type="free"), nperm=999)

set.seed(1)
adonis2(t(otu_table(ps_reduced))~age_gr+study_gr, data=as.data.frame(as.matrix(sample_data(ps_reduced))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(t(otu_table(ps_reduced))~age_gr+study_gr, data=as.data.frame(as.matrix(sample_data(ps_reduced))), method="jaccard", permutations=perm, by="margin", binary=T)

#Determine the effect of medicine for patients with older diagnosis:
perm <- how(within=Within(type="free"), plots=Plots(strata=sample_data(ps_old)$person_ID, type="none"), nperm=999)

set.seed(1)
adonis2(t(otu_table(ps_old))~bio+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(t(otu_table(ps_old))~bio+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="jaccard", permutations=perm, by="margin", binary=T)

set.seed(1)
adonis2(t(otu_table(ps_old))~aza+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(t(otu_table(ps_old))~aza+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="jaccard", permutations=perm, by="margin", binary=T)

set.seed(1)
adonis2(t(otu_table(ps_old))~pred+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(t(otu_table(ps_old))~pred+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="jaccard", permutations=perm, by="margin", binary=T)

set.seed(1)
adonis2(t(otu_table(ps_old))~asa+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(t(otu_table(ps_old))~asa+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="jaccard", permutations=perm, by="margin", binary=T)
```


# Stability analysis of the microbiome across age groups and study groups

Investigation of the flucturation of the microbiome. Is it depending on age group? Or on study group?
Calculation of distance (beta diversity) between sample 1 and sample 5 from the same individual. Is this internal distance depending on the groups?
```{r}
#Select timepoint 1 and timepoint 5
df <- data.frame(sample_data(ps_relab))
df <- df %>% filter(time_point %in% c(0,4))
n_df <- df %>% group_by(person_ID) %>% dplyr::summarise(n=n()) %>% filter(n>1) 
t1t5 <- df %>% filter(person_ID %in% n_df$person_ID)
ps_t1t5 <- subset_samples(ps_relab, J_ID %in% t1t5$J_ID)

#Calculate beta div
bc_meta <-  phyloseq::distance(ps_t1t5, method="bray")
jac_meta <- phyloseq::distance(ps_t1t5, method="jaccard", binary=T)

#Transform into table
bc_meta <- as.matrix(bc_meta)
bc_meta <- as.data.frame(bc_meta)
jac_meta <- as.matrix(jac_meta)
jac_meta <- as.data.frame(jac_meta)

#Make new dataframe
t1 <- t1t5 %>% group_by(person_ID) %>% arrange(time_point) %>% slice(1) %>% ungroup()
t5 <- t1t5 %>% group_by(person_ID) %>% arrange(time_point) %>% slice(n()) %>% ungroup()

df_dist<- tibble(person_ID=t1$person_ID, J1 = t1$J_ID, J2=t5$J_ID, dist_bc=rep(NA, nrow(t1)), dist_jac=rep(NA, nrow(t1)),age_gr= as.factor(t1$age_gr), study_gr= as.factor(t1$study_gr))

df_dist <- as.data.frame(df_dist)
for (i in 1:nrow(df_dist)){
  df_dist[i,"dist_bc"] <- bc_meta[df_dist$J1[i], df_dist$J2[i]]
  df_dist[i,"dist_jac"] <- jac_meta[df_dist$J1[i], df_dist$J2[i]]
}

#Make plots:
df_dist$study_gr1 <- ifelse(df_dist$study_gr==1, "Newly diagnosed", ifelse(df_dist$study_gr==2, "Older diagnosis", NA))

p1 <- ggplot(data=df_dist,aes(x=study_gr1, y=dist_bc))+
  geom_violin(aes(color=study_gr1, fill=study_gr1))+geom_dotplot(binaxis='y', stackdir='center', dotsize=1)+theme_classic()+theme(legend.position = "none", text = element_text(family = "sans", size = 8))+ xlab(" ") + ylab("Bray Curtis dissimilarity")

p2 <- ggplot(data=df_dist,aes(x=study_gr1, y=dist_jac))+
  geom_violin(aes(color=study_gr1, fill=study_gr1))+geom_dotplot(binaxis='y', stackdir='center', dotsize=1)+theme_classic()+theme(legend.position = "none", text = element_text(family = "sans", size = 8))+ xlab(" ") + ylab("Jaccard distance")

ggarrange(p1, p2)

df_dist$age_gr1 <- ifelse(df_dist$age_gr==1, "Children", ifelse(df_dist$age_gr==2, "Adults", NA))
df_dist$age_gr1 <- factor(df_dist$age_gr1, levels=c("Children", "Adults"))

p3 <- ggplot(data=df_dist,aes(x=age_gr1, y=dist_bc))+
  geom_violin(aes( color=age_gr1, fill=age_gr1))+geom_dotplot(binaxis='y', stackdir='center', dotsize=1)+theme_classic()+theme(legend.position = "none", text = element_text(family = "sans", size = 8))+
  xlab(" ") + ylab("Bray Curtis dissimilarity")

p4 <- ggplot(data=df_dist,aes(x=age_gr1, y=dist_jac))+
  geom_violin(aes( color=age_gr1, fill=age_gr1))+geom_dotplot(binaxis='y', stackdir='center', dotsize=1)+theme_classic()+theme(legend.position = "none", text = element_text(family = "sans", size = 8))+
  xlab(" ") + ylab("Jaccard distance")
ggarrange(p3, p4)

#Gather figure supp 2
p_top <- ggarrange(p1, p2, ncol=2, labels = c("A)", "C)"), font.label = list(size=10, family="sans"), widths = c(1, 1))

p_bottom <- ggarrange(p3, p4, labels=c("B)", "D)"), font.label = list(size=10, family="sans"), widths = c(1, 1))

p_total <- ggarrange(p_top, p_bottom, nrow=2, heights = c(1,1))

ggsave(p_total, filename = "C:/Users/FX76TZ/OneDrive - Aalborg Universitet/IBD project/V1/illustrations/Manuscript/Figure_Supp2.tiff", width = 174, height = 160, units = "mm", dpi=300)


#Divide based on study group
df_new <- df_dist %>% filter(study_gr == 1)
df_old <- df_dist %>% filter(study_gr == 2)
#Test if there is a difference using permutation ANOVA: Bray Curtis
set.seed(1)
res <- anova(lm(formula=dist_bc~age_gr, data=df_new))
res
summary(lm(formula=dist_bc~age_gr, data=df_new))

df_test_age <- rep(NA, 1000)
for (i in 1:1000){
    #Permute and save F-statistic: Age group
    df_shuffle_age <- df_new
    df_shuffle_age$age_gr <- sample(x=df_shuffle_age$age_gr, size=nrow(df_new))
    test_age <- as.data.frame(anova(lm(formula=dist_bc~age_gr, data=df_shuffle_age)))
    df_test_age[i] <- test_age["age_gr", "F value"]
  }

#Make p-value
res_age <- res["age_gr", "F value"]
sum(df_test_age>=res_age)/1000

#Jaccard:
set.seed(1)
res <- anova(lm(formula=dist_jac~age_gr, data=df_new))
res
summary(lm(formula=dist_jac~age_gr, data=df_new))

df_test_age <- rep(NA, 1000)
for (i in 1:1000){
    #Permute and save F-statistic: Age group
    df_shuffle_age <- df_new
    df_shuffle_age$age_gr <- sample(x=df_shuffle_age$age_gr, size=nrow(df_new))
    test_age <- as.data.frame(anova(lm(formula=dist_jac~age_gr, data=df_shuffle_age)))
    df_test_age[i] <- test_age["age_gr", "F value"]
  }

#Make p-value
res_age <- res["age_gr", "F value"]
sum(df_test_age>=res_age)/1000

#Repeat for ealier diagnosed:
set.seed(1)
res <- anova(lm(formula=dist_bc~age_gr, data=df_old))
res
summary(lm(formula=dist_bc~age_gr, data=df_old))

df_test_age <- rep(NA, 1000)
for (i in 1:1000){
    #Permute and save F-statistic: Age group
    df_shuffle_age <- df_old
    df_shuffle_age$age_gr <- sample(x=df_shuffle_age$age_gr, size=nrow(df_old))
    test_age <- as.data.frame(anova(lm(formula=dist_bc~age_gr, data=df_shuffle_age)))
    df_test_age[i] <- test_age["age_gr", "F value"]
  }

#Make p-value
res_age <- res["age_gr", "F value"]
sum(df_test_age>=res_age)/1000

#Jaccard:
set.seed(1)
res <- anova(lm(formula=dist_jac~age_gr, data=df_old))
res
summary(lm(formula=dist_jac~age_gr, data=df_old))

df_test_age <- rep(NA, 1000)
for (i in 1:1000){
    #Permute and save F-statistic: Age group
    df_shuffle_age <- df_old
    df_shuffle_age$age_gr <- sample(x=df_shuffle_age$age_gr, size=nrow(df_old))
    test_age <- as.data.frame(anova(lm(formula=dist_jac~age_gr, data=df_shuffle_age)))
    df_test_age[i] <- test_age["age_gr", "F value"]
  }

#Make p-value
res_age <- res["age_gr", "F value"]
sum(df_test_age>=res_age)/1000

#Repeat for the whole dataset:
set.seed(1)
res <- anova(lm(formula=dist_bc~age_gr+study_gr, data=df_dist))
res
summary(lm(formula=dist_bc~age_gr+study_gr, data=df_dist))

df_test_age <- rep(NA, 1000)
for (i in 1:1000){
    #Permute and save F-statistic: Age group
    df_shuffle_age <- df_dist
    df_shuffle_age$age_gr <- sample(x=df_shuffle_age$age_gr, size=nrow(df_dist))
    test_age <- as.data.frame(anova(lm(formula=dist_bc~age_gr+study_gr, data=df_shuffle_age)))
    df_test_age[i] <- test_age["age_gr", "F value"]
  }

#Make p-value
res_age <- res["age_gr", "F value"]
sum(df_test_age>=res_age)/1000

#Jaccard:
set.seed(1)
res <- anova(lm(formula=dist_jac~age_gr+study_gr, data=df_dist))
res
summary(lm(formula=dist_jac~age_gr+study_gr, data=df_dist))

df_test_age <- rep(NA, 1000)
for (i in 1:1000){
    #Permute and save F-statistic: Age group
    df_shuffle_age <- df_dist
    df_shuffle_age$age_gr <- sample(x=df_shuffle_age$age_gr, size=nrow(df_dist))
    test_age <- as.data.frame(anova(lm(formula=dist_jac~age_gr+study_gr, data=df_shuffle_age)))
    df_test_age[i] <- test_age["age_gr", "F value"]
  }

#Make p-value
res_age <- res["age_gr", "F value"]
sum(df_test_age>=res_age)/1000

```

