---
title: "Analyses of single species using NBZIMM and ZIGMM"
output:
  github_document:
    toc: true
editor_options:
  chunk_output_type: console
---
# Read in packages and data

Read in packages and path to folder:
```{r echo=T, message=F, error=F, warning=F}
library(tidyverse)
library(caret)
library(phyloseq)
library(NBZIMM)
library(vegan)
library(metaMint)
library(lme4)
library(lmerTest)
library(zCompositions)
path = "C:/Users/Tom/Documents/10. semester/V1/data/"

```

Read in data:
```{r}
#Phyloseq objects:
ps_count <- readRDS(paste0(path, "humann3_processed_output/Phyloseq/","ps_count_50mean_50var_zero.rds"))

#Meta data:
meta <- read.delim(paste0(path, "metadata/", "preped_metadata_filtered.txt"))

#Merge meta data with phyloseq object to remove those samples with >85% reads removed as human 
rownames(meta) <- meta$J_ID
ps_count <- phyloseq(otu_table(ps_count), tax_table(ps_count), sample_data(meta))
```

# Test the association with clinical disease score using NBZIMM

Fit the NBZIMM: Test the clinical disease score
```{r}
#Analysis of MetaPhlAn count data - all taxa!:
ps_count <- subset_samples(ps_count, !is.na(score))
N <- sample_data(ps_count)$rdepth
person_ID <- as.factor(sample_data(ps_count)$person_ID)
sex <-  as.factor(sample_data(ps_count)$sex)
study_gr <-  as.factor(sample_data(ps_count)$study_gr)
age_gr <-  as.factor(sample_data(ps_count)$age_gr)
asa <-  as.factor(sample_data(ps_count)$asa)
bio <-  as.factor(sample_data(ps_count)$bio)
pred <-  as.factor(sample_data(ps_count)$pred_total)
score_num <- sample_data(ps_count)$score_num
time_point <- sample_data(ps_count)$time_point

clinical = cbind.data.frame(N, person_ID, sex,study_gr,age_gr, asa, bio, pred, score_num, time_point)
pheno <-  as.data.frame(as.matrix(t(otu_table(ps_count)))) #No transformation

#Fit model with discrete numeric and binary disease score. Fit with interaction with age:
f_inter <- mms(y=pheno, fixed= ~study_gr+sex+asa+bio+pred+age_gr*score_num + offset(log(N)), data=clinical, random = ~1|person_ID, method="zinb", correlation=corAR1(form=~time_point|person_ID), zi_fixed=~study_gr+sex+asa+bio+pred+age_gr*score_num + offset(log(N)))

failed <- c(8,10,12,18,23,24,30,70,72,75,76,77)
failed <- colnames(pheno)[failed]

#Model those that failed as relative abundancen instead,   
pheno_relab <- pheno
for (i in 1:dim(pheno)[1]){
  pheno_relab[i, ] <- pheno_relab[i, ] / N[i]
}

#Is it gaussian?
for (i in colnames(pheno_relab)[1:5]){
  p <- ggplot(pheno_relab)+
    geom_histogram(aes_string(x=i), bins=30)+ggtitle(i)
  print(p)
}

for (i in colnames(pheno_relab)[1:5]){
  test <- asin(sqrt(pheno_relab[, i]))
  p <- ggplot()+
    geom_histogram(aes_string(x=test), bins=30)+ggtitle(i)
  print(p)
}
#It is better with arcsine square root transformation

#ZIG - with asin+sqrt
f_inter_relab <- mms(y=asin(sqrt(pheno_relab[, failed])), fixed= ~study_gr+sex+asa+bio+pred+age_gr*score_num, data=clinical, random = ~1|person_ID, method="zig", correlation=corAR1(form=~time_point|person_ID), zi_fixed=~study_gr+sex+asa+bio+pred+age_gr*score_num)


#Find taxa with significant (p-adj<0.05) interaction.  
res_inter=as.data.frame(get.fixed(f_inter, part="dist", vr.name="age_gr2:score_num"))
res_inter$bugs <- rownames(res_inter)
res_inter$method <- "NBZIMM"

res_inter_relab <- as.data.frame(get.fixed(f_inter_relab, part="dist", vr.name="age_gr2:score_num"))
res_inter_relab$bugs <- rownames(res_inter_relab)
res_inter_relab$method <- "ZIGMM"

#Merge tables to calculate p_adj for all:
res_total <- rbind(res_inter, res_inter_relab)
res_total$p_adj <- p.adjust(res_total$pvalue, method="BH")
write.table(res_total, file = paste0(path, "Results/", "Clinical_interaction.txt"), col.names=T, row.names=F)
res_inter_sig <- res_total %>% filter(p_adj < 0.05)
res_inter <- res_total

##Select the rest of those not significant and fit without the interaction term:
pheno_1 <- pheno[, setdiff(colnames(pheno), rownames(res_inter_sig))]

f_nointer<- mms(y=pheno_1, fixed= ~study_gr+sex+asa+bio+pred+age_gr+score_num + offset(log(N)), data=clinical, random = ~1|person_ID, method="zinb", correlation=corAR1(form=~time_point|person_ID), zi_fixed=~study_gr+sex+asa+bio+pred+age_gr+score_num + offset(log(N)))
failed2 <- c(4,7,8,10,11,19,25,26,47,64,66,68,70)
failed2 <- colnames(pheno_1)[failed2]

#Fit using ZIGMM:
f_nointer_relab <- mms(y=asin(sqrt(pheno_relab[, failed2])), fixed= ~study_gr+sex+asa+bio+pred+age_gr+score_num, data=clinical, random = ~1|person_ID, method="zig", correlation=corAR1(form=~time_point|person_ID), zi_fixed=~study_gr+sex+asa+bio+pred+age_gr+score_num)

#Result: Disease score in interaction:
res=as.data.frame(get.fixed(f_inter, part="dist", vr.name="score_num"))
res$bugs <- rownames(res)
res$method <- "NBZIMM"

res_relab=as.data.frame(get.fixed(f_inter_relab, part="dist", vr.name="score_num"))
res_relab$bugs <- rownames(res_relab)
res_relab$method <- "ZIGMM"

res_total <- rbind(res, res_relab)
res_total$p_adj <- p.adjust(res_total$pvalue, method="BH")

write.table(res_total, file = paste0(path, "Results/", "Score_interaction_disease.txt"), col.names=T, row.names=F)
res_inter_disease <- res_total

#Result: Interaction in zero-inflated part:
res=as.data.frame(get.fixed(f_inter, part="zero", vr.name="age_gr2:score_num"))
res$bugs <- rownames(res)
res$method <- "NBZIMM"

res_relab=as.data.frame(get.fixed(f_inter_relab, part="zero", vr.name="age_gr2:score_num"))
res_relab$bugs <- rownames(res_relab)
res_relab$method <- "ZIGMM"

res_total <- rbind(res, res_relab)
res_total$p_adj <- p.adjust(res_total$pvalue, method="BH")

write.table(res_total, file = paste0(path, "Results/", "Score_interaction_zero.txt"), col.names=T, row.names=F)
res_inter_zero <- res_total

#Result: Disease score in interaction, zero-inflated:
res=as.data.frame(get.fixed(f_inter, part="zero", vr.name="score_num"))
res$bugs <- rownames(res)
res$method <- "NBZIMM"

res_relab=as.data.frame(get.fixed(f_inter_relab, part="zero", vr.name="score_num"))
res_relab$bugs <- rownames(res_relab)
res_relab$method <- "ZIGMM"

res_total <- rbind(res, res_relab)
res_total$p_adj <- p.adjust(res_total$pvalue, method="BH")

write.table(res_total, file = paste0(path, "Results/", "Score_interaction_zero_disease.txt"), col.names=T, row.names=F)
res_inter_zero_disease <- res_total

#Result: Disease score without interaction:
res=as.data.frame(get.fixed(f_nointer, part="dist", vr.name="score_num"))
res$bugs <- rownames(res)
res$method <- "NBZIMM"

res_relab=as.data.frame(get.fixed(f_nointer_relab, part="dist", vr.name="score_num"))
res_relab$bugs <- rownames(res_relab)
res_relab$method <- "ZIGMM"

res_total <- rbind(res, res_relab)
res_total$p_adj <- p.adjust(res_total$pvalue, method="BH")

write.table(res_total, file = paste0(path, "Results/", "Score_disease.txt"), col.names=T, row.names=F)
res_disease <- res_total

#Result: Disease score without interaction, zero-inflated:
res=as.data.frame(get.fixed(f_nointer, part="zero", vr.name="score_num"))
res$bugs <- rownames(res)
res$method <- "NBZIMM"

res_relab=as.data.frame(get.fixed(f_nointer_relab, part="zero", vr.name="score_num"))
res_relab$bugs <- rownames(res_relab)
res_relab$method <- "ZIGMM"

res_total <- rbind(res, res_relab)
res_total$p_adj <- p.adjust(res_total$pvalue, method="BH")

write.table(res_total, file = paste0(path, "Results/", "Score_zero_disease.txt"), col.names=T, row.names=F)
res_disease_zero <- res_total

```

# Test the association with clinical disease score using clr transformation and ZIGMM

Perform the same analyses but with clr transformed microbiome data, to compare the two methods:
```{r}
#Different clr transformation: mclr. It keeps the zeroes as they are, calculate clr from the nonzero data and shifts the clr transformed data to positive values. 
#Analysis of MetaPhlAn count data - all taxa!:
pheno_clr <-  as.matrix(t(otu_table(ps_count)))
pheno_clr <- mclr(pheno_clr)

#Fit using ZIGMM with interaction term:
f_inter_clr <- mms(y=pheno_clr, fixed= ~study_gr+sex+asa+bio+pred+age_gr*score_num, data=clinical, random = ~1|person_ID, method="zig", correlation=corAR1(form=~time_point|person_ID), zi_fixed=~study_gr+sex+asa+bio+pred+age_gr*score_num)

#Find taxa with significant (p-adj<0.05) interaction.  
res_inter_clr=as.data.frame(get.fixed(f_inter_clr, part="dist", vr.name="age_gr2:score_num"))
res_inter_clr$bugs <- rownames(res_inter_clr)
res_inter_clr$method <- "CLR-ZIGMM"
res_inter_clr$p_adj <- p.adjust(res_inter_clr$pvalue, method = "BH")
write.table(res_inter_clr, file = paste0(path, "Results/", "CLR_Clinical_interaction.txt"), col.names=T, row.names=F)

res_inter_clr_sig <- res_inter_clr %>% filter(p_adj < 0.05)

##Select the rest of those not significant and fit without the interaction term:
pheno_clr_1 <- pheno_clr[, setdiff(colnames(pheno_clr), rownames(res_inter_clr_sig))]

f_nointer_clr<- mms(y=pheno_clr_1, fixed= ~study_gr+sex+asa+bio+pred+age_gr+score_num, data=clinical, random = ~1|person_ID, method="zig", correlation=corAR1(form=~time_point|person_ID), zi_fixed=~study_gr+sex+asa+bio+pred+age_gr+score_num)


#Result: Disease score in interaction:
res=as.data.frame(get.fixed(f_inter_clr, part="dist", vr.name="score_num"))
res$bugs <- rownames(res)
res$method <- "CLR-ZIGMM"
res$p_adj <- p.adjust(res$pvalue, method="BH")

write.table(res, file = paste0(path, "Results/", "CLR_Score_interaction_disease.txt"), col.names=T, row.names=F)
res_inter_disease_clr <- res

#Result: Interaction in zero-inflated part:
res=as.data.frame(get.fixed(f_inter_clr, part="zero", vr.name="age_gr2:score_num"))
res$bugs <- rownames(res)
res$method <- "CLR-ZIGMM"
res$p_adj <- p.adjust(res$pvalue, method="BH")

write.table(res, file = paste0(path, "Results/", "CLR_Score_interaction_zero.txt"), col.names=T, row.names=F)
res_inter_zero_clr <- res

#Result: Disease score in interaction, zero-inflated:
res=as.data.frame(get.fixed(f_inter_clr, part="zero", vr.name="score_num"))
res$bugs <- rownames(res)
res$method <- "CLR-ZIGMM"
res$p_adj <- p.adjust(res$pvalue, method="BH")

write.table(res, file = paste0(path, "Results/", "CLR_Score_interaction_zero_disease.txt"), col.names=T, row.names=F)
res_inter_zero_disease_clr <- res

#Result: Disease score without interaction:
res=as.data.frame(get.fixed(f_nointer_clr, part="dist", vr.name="score_num"))
res$bugs <- rownames(res)
res$method <- "CLR-ZIGMM"
res$p_adj <- p.adjust(res$pvalue, method="BH")

write.table(res, file = paste0(path, "Results/", "CLR_Score_disease.txt"), col.names=T, row.names=F)
res_disease_clr <- res

#Result: Disease score without interaction, zero-inflated:
res=as.data.frame(get.fixed(f_nointer_clr, part="zero", vr.name="score_num"))
res$bugs <- rownames(res)
res$method <- "CLR-ZIGMM"
res$p_adj <- p.adjust(res$pvalue, method="BH")

write.table(res, file = paste0(path, "Results/", "CLR_Score_zero_disease.txt"), col.names=T, row.names=F)
res_disease_zero_clr <- res
```

# Evaluate results from clinical disease score association analyses 

Investigate the results from the count and relab data: 
```{r}
#Make histograms of p-values from the interaction analysis:
ggplot(res_inter)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Interaction term")


#Extract the significant in interaction term together with the disease score estimate
sig <- res_inter %>% filter(p_adj<0.05)
sig
sig1 <- res_inter_disease[row.names(sig),]
sig1
sig_inter <- sig

#Plot significant interaction:
#Plot the detected species - interaction:
pheno_temp <- pheno
pheno_temp$J_ID <- row.names(pheno_temp)
temp_meta <- merge(pheno_temp, meta, by.x="J_ID", by.y="J_ID")
temp_meta$score_num <- as.factor(temp_meta$score_num)
temp_meta$age_gr1 <- ifelse(temp_meta$age_gr==1, "Children", "Adults")
temp_meta$age_gr1 <-factor(temp_meta$age_gr1, levels=c("Children", "Adults"))
temp_meta$score_num1 <- ifelse(temp_meta$score_num==1, "Remission", ifelse(temp_meta$score_num==2, "Mild", ifelse(temp_meta$score_num==3, "Moderate", "Severe")))
temp_meta$score_num1 <-factor(temp_meta$score_num1, levels=c("Remission", "Mild", "Moderate", "Severe"))

spec <- sig$bugs
for (i in spec){
  j <- gsub("s__", "", i)
  j <- gsub("_", " ", j)
  p <- ggplot(data=temp_meta)+
    geom_boxplot(aes_string(x="score_num1", y=paste0(i, "/rdepth"), color = "age_gr1"), outlier.size=1)+theme_classic()+xlab("Clinical disease score")+ylab("Relative abundance") + ggtitle(as.character(j))+labs(color="Age group")
 print(p)
}

#Plot all NBZIMM interaction results together in one plot:
pheno_temp_relab <- pheno_relab
pheno_temp_relab$J_ID <- row.names(pheno_temp_relab)
temp_meta_relab <- merge(pheno_temp_relab, meta, by.x="J_ID", by.y="J_ID")
temp_meta_relab$score_num <- as.factor(temp_meta_relab$score_num)
temp_meta_relab$age_gr1 <- ifelse(temp_meta_relab$age_gr==1, "Children", "Adults")
temp_meta_relab$age_gr1 <-factor(temp_meta_relab$age_gr1, levels=c("Children", "Adults"))
temp_meta_relab$score_num1 <- ifelse(temp_meta_relab$score_num==1, "Remission", ifelse(temp_meta_relab$score_num==2, "Mild", ifelse(temp_meta_relab$score_num==3, "Moderate", "Severe")))
temp_meta_relab$score_num1 <-factor(temp_meta_relab$score_num1, levels=c("Remission", "Mild", "Moderate", "Severe"))
temp1 <- temp_meta_relab %>%
  pivot_longer(., cols = colnames(pheno_relab), names_to = "Var", values_to = "Val")

temp1$Var1 <- gsub("s__", "", temp1$Var)
temp1$Var1 <- gsub("_", " ", temp1$Var1)

temp2 <- temp1 %>% filter(Var %in% spec)
ggplot(temp2, aes(x=score_num1, y=Val))+
  geom_boxplot(aes(color=age_gr1))+theme_classic()+facet_wrap(~Var1, scales="free_y")+ylab("Relative abundance")+labs(color="Age group")+ theme(legend.position = c(1, 0),legend.justification = c(1, 0))+xlab("")

#Extract the significant in interaction term in the zero-inflated part, together with the disease score estimate
ggplot(res_inter_zero)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Zero-inflated interaction term")

sig <- res_inter_zero %>% filter(p_adj<0.05)
sig #None


#Investigate results without interaction term: Disease score:
ggplot(res_disease)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Disease score")

sig <- res_disease %>% filter(p_adj<0.05)
sig
sig_disease <- sig

#Plot the detected species:
spec <- sig$bugs
for (i in spec){
  j <- gsub("s__", "", i)
  j <- gsub("_", " ", j)
  p <- ggplot(data=temp_meta)+
    geom_jitter(aes_string(x="score_num1", y=paste0(i, "/rdepth"), color = "score_num1"), size=1)+theme_classic()+xlab("Clinical disease score")+ylab("Relative abundance") + ggtitle(as.character(j))+theme(legend.position = "none")
 print(p)
}
#In one plot:
temp2 <- temp1 %>% filter(Var %in% spec)
ggplot(data=temp2)+
    geom_jitter(aes(x=score_num1, y=Val, color = score_num1), size=1)+theme_classic()+facet_wrap(~Var1, scales="free_y", ncol=3)+xlab("Clinical disease score")+ylab("Relative abundance")+theme(legend.position = "none")


#Investigate results without interaction term: Disease score, zero inflated:
ggplot(res_disease_zero)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Zero-inflated disease score")

sig <- res_disease_zero %>% filter(p_adj<0.05)
sig
sig_disease_zero <- sig

#Plot the results:
#Plot the detected species:
spec <- sig$bugs
for (i in spec){
  j <- gsub("s__", "", i)
  j <- gsub("_", " ", j)
  test <- temp_meta[,i]
  test <- ifelse(test==0, "Absent", "Present")
  temp_meta$test <- test
  p <- ggplot(data=temp_meta)+
    geom_bar(aes(x=score_num1, fill=test), position="fill")+theme_classic()+xlab("Clinical disease score")+ylab(" ") + ggtitle(as.character(j))+labs(fill="")
print(p)

}

```

Visualize the CLR results and compare!
```{r}

#Make histograms of p-values from the interaction analysis:
ggplot(res_inter_clr)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Interaction term")


#Extract the significant in interaction term together with the disease score estimate
sig <- res_inter_clr %>% filter(p_adj<0.05)
sig 
sig_inter_clr <- sig

#Extract the significant in interaction term in the zero-inflated part, together with the disease score estimate
ggplot(res_inter_zero_clr)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Zero-inflated interaction term")

sig <- res_inter_zero_clr %>% filter(p_adj<0.05)
sig
sig1 <- res_inter_zero_disease_clr[row.names(sig),]
sig1
sig_inter_zero_clr <- sig

#Plot the results:
#Plot the detected species:
spec <- sig$bugs
for (i in spec){
  test <- temp_meta[,i]
  test <- ifelse(test==0, "Absent", "Present")
  temp_meta$test <- test
  p <- ggplot(data=temp_meta)+
    geom_bar(aes(x=score_num, fill=test), position="fill")+theme_classic()+xlab("Disease score")+ylab(" ") + ggtitle(as.character(i))+labs(fill="")+facet_wrap(~age_gr)
print(p)

}

#Investigate results without interaction term: Disease score:
ggplot(res_disease_clr)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Disease score")

sig <- res_disease_clr %>% filter(p_adj<0.05)
sig
sig_disease_clr <- sig

#Plot the detected species:
spec <- sig$bugs
for (i in spec){
  p <- ggplot(data=temp_meta)+
    geom_jitter(aes_string(x="score_num", y=paste0(i, "/rdepth"), color = "score_num"))+theme_classic()+xlab("Disease score")+ylab("Relative abundance") + ggtitle(as.character(i))+theme(legend.position = "none")
print(p)

}


#Investigate results without interaction term: Disease score, zero inflated:
ggplot(res_disease_zero_clr)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Zero-inflated disease score")

sig <- res_disease_zero_clr %>% filter(p_adj<0.05)
sig #None!
```

Compare the two methods:
```{r}
sig_inter$test <- "Interaction: NBZIMM, NB"
sig_inter_clr$test <- "Interaction: ZIGMM, G"
sig_inter_zero_clr$test <- "Interaction: ZIGMM, ZI"
sig_disease$test <- "Disease score: NBZIMM, NB"
sig_disease_clr$test <- "Disease score: ZIGMM, G"
sig_disease_zero$test <-  "Disease score: NBZIMM, ZI"

#Transform estimates to enable comparison:Transformed to model presence instead of absence of species
sig_inter_zero_clr$Estimate <- -sig_inter_zero_clr$Estimate
sig_disease_zero$Estimate <- -sig_disease_zero$Estimate

res_total <- rbind(sig_inter,sig_inter_clr, sig_inter_zero_clr, sig_disease, sig_disease_zero, sig_disease_clr)
res_total$test <- factor(res_total$test, levels=c("Interaction: NBZIMM, NB","Interaction: ZIGMM, G","Interaction: ZIGMM, ZI", "Disease score: NBZIMM, NB","Disease score: NBZIMM, ZI", "Disease score: ZIGMM, G"))

res_total$bugs <- gsub("s__", "", res_total$bugs)
res_total$bugs <- gsub("_", " ", res_total$bugs)

bl <- colorRampPalette(c("navy","royalblue","lightskyblue"))(200)   
re <- colorRampPalette(c("mistyrose", "red2","darkred"))(200)

min(res_total$Estimate)
max(res_total$Estimate)

ggplot(data=res_total)+
  geom_tile(aes(x=test, y=bugs, fill=Estimate))+ylab("")+xlab(" ")+
  scale_fill_gradientn(colours=c(bl, re), na.value = "black",
         limits = c(-2.6, 2.6))+theme_classic()+theme(axis.text.x = element_text(angle = 90))


```

# Test the association with paraclinical disease score using NBZIMM 

Test everything again, but with faecal calprotectin as disease score: 
Read in data:
```{r}
#Phyloseq objects:
ps_count <- readRDS(paste0(path, "humann3_processed_output/Phyloseq/","ps_count_50mean_50var_zero.rds"))

#Meta data:
meta <- read.delim(paste0(path, "metadata/", "preped_metadata_filtered.txt"))

#Merge meta data with phyloseq object to remove those samples with >85% reads removed as human 
rownames(meta) <- meta$J_ID
ps_count <- phyloseq(otu_table(ps_count), tax_table(ps_count), sample_data(meta))
```

Fit the NBZIMM: Test the para-clinical disease score
```{r}
#Analysis of MetaPhlAn count data - all taxa!:
ps_count <- subset_samples(ps_count, !is.na(f_cal_current))
N <- sample_data(ps_count)$rdepth
person_ID <- as.factor(sample_data(ps_count)$person_ID)
sex <-  as.factor(sample_data(ps_count)$sex)
study_gr <-  as.factor(sample_data(ps_count)$study_gr)
age_gr <-  as.factor(sample_data(ps_count)$age_gr)
asa <-  as.factor(sample_data(ps_count)$asa)
bio <-  as.factor(sample_data(ps_count)$bio)
pred <-  as.factor(sample_data(ps_count)$pred_total)
fcal <- log(sample_data(ps_count)$f_cal_current+1)
time_point <- sample_data(ps_count)$time_point

clinical = cbind.data.frame(N, person_ID, sex,study_gr,age_gr, asa, bio, pred, fcal, time_point)
pheno <-  as.data.frame(as.matrix(t(otu_table(ps_count)))) #No transformation

#Fit model with discrete numeric and binary disease score. Fit with interaction with age:
f_inter <- mms(y=pheno, fixed= ~study_gr+sex+asa+bio+pred+age_gr*fcal + offset(log(N)), data=clinical, random = ~1|person_ID, method="zinb", correlation=corAR1(form=~time_point|person_ID), zi_fixed=~study_gr+sex+asa+bio+pred+age_gr*fcal + offset(log(N)))

failed <- c(6,8,10,12,18,22,30,48,53,70,72,74,75,76)
failed <- colnames(pheno)[failed]

#Model those that failed as relative abundancen instead,   
pheno_relab <- pheno
for (i in 1:dim(pheno)[1]){
  pheno_relab[i, ] <- pheno_relab[i, ] / N[i]
}


#ZIG - with asin+sqrt
f_inter_relab <- mms(y=asin(sqrt(pheno_relab[, failed])), fixed= ~study_gr+sex+asa+bio+pred+age_gr*fcal, data=clinical, random = ~1|person_ID, method="zig", correlation=corAR1(form=~time_point|person_ID), zi_fixed=~study_gr+sex+asa+bio+pred+age_gr*fcal)

#Find taxa with significant (p-adj<0.05) interaction.  
res_inter=as.data.frame(get.fixed(f_inter, part="dist", vr.name="age_gr2:fcal"))
res_inter$bugs <- rownames(res_inter)
res_inter$method <- "NBZIMM"

res_inter_relab <- as.data.frame(get.fixed(f_inter_relab, part="dist", vr.name="age_gr2:fcal"))
res_inter_relab$bugs <- rownames(res_inter_relab)
res_inter_relab$method <- "ZIGMM"

#Merge tables to calculate p_adj for all:
res_total <- rbind(res_inter, res_inter_relab)
res_total$p_adj <- p.adjust(res_total$pvalue, method="BH")
write.table(res_total, file = paste0(path, "Results/", "ParaClinical_interaction.txt"), col.names=T, row.names=F)
res_inter_sig <- res_total %>% filter(p_adj < 0.05)
res_inter <- res_total

##Select the rest of those not significant and fit without the interaction term:
pheno_1 <- pheno[, setdiff(colnames(pheno), rownames(res_inter_sig))]

f_nointer<- mms(y=pheno_1, fixed= ~study_gr+sex+asa+bio+pred+age_gr+fcal + offset(log(N)), data=clinical, random = ~1|person_ID, method="zinb", correlation=corAR1(form=~time_point|person_ID), zi_fixed=~study_gr+sex+asa+bio+pred+age_gr+fcal + offset(log(N)))
failed2 <- c(4,6,8,9,11,20,28,45,50,52,66,67,69,70,71)
failed2 <- colnames(pheno_1)[failed2]

#Fit using ZIGMM:
f_nointer_relab <- mms(y=asin(sqrt(pheno_relab[, failed2])), fixed= ~study_gr+sex+asa+bio+pred+age_gr+fcal, data=clinical, random = ~1|person_ID, method="zig", correlation=corAR1(form=~time_point|person_ID), zi_fixed=~study_gr+sex+asa+bio+pred+age_gr+fcal)

#Result: Disease score in interaction:
res=as.data.frame(get.fixed(f_inter, part="dist", vr.name="fcal"))
res$bugs <- rownames(res)
res$method <- "NBZIMM"

res_relab=as.data.frame(get.fixed(f_inter_relab, part="dist", vr.name="fcal"))
res_relab$bugs <- rownames(res_relab)
res_relab$method <- "ZIGMM"

res_total <- rbind(res, res_relab)
res_total$p_adj <- p.adjust(res_total$pvalue, method="BH")

write.table(res_total, file = paste0(path, "Results/", "Paraclinical_interaction_disease.txt"), col.names=T, row.names=F)
res_inter_disease <- res_total

#Result: Interaction in zero-inflated part:
res=as.data.frame(get.fixed(f_inter, part="zero", vr.name="age_gr2:fcal"))
res$bugs <- rownames(res)
res$method <- "NBZIMM"

res_relab=as.data.frame(get.fixed(f_inter_relab, part="zero", vr.name="age_gr2:fcal"))
res_relab$bugs <- rownames(res_relab)
res_relab$method <- "ZIGMM"

res_total <- rbind(res, res_relab)
res_total$p_adj <- p.adjust(res_total$pvalue, method="BH")

write.table(res_total, file = paste0(path, "Results/", "Paraclinical_interaction_zero.txt"), col.names=T, row.names=F)
res_inter_zero <- res_total

#Result: Disease score in interaction, zero-inflated:
res=as.data.frame(get.fixed(f_inter, part="zero", vr.name="fcal"))
res$bugs <- rownames(res)
res$method <- "NBZIMM"

res_relab=as.data.frame(get.fixed(f_inter_relab, part="zero", vr.name="fcal"))
res_relab$bugs <- rownames(res_relab)
res_relab$method <- "ZIGMM"

res_total <- rbind(res, res_relab)
res_total$p_adj <- p.adjust(res_total$pvalue, method="BH")

write.table(res_total, file = paste0(path, "Results/", "Paraclinical_interaction_zero_disease.txt"), col.names=T, row.names=F)
res_inter_zero_disease <- res_total

#Result: Disease score without interaction:
res=as.data.frame(get.fixed(f_nointer, part="dist", vr.name="fcal"))
res$bugs <- rownames(res)
res$method <- "NBZIMM"

res_relab=as.data.frame(get.fixed(f_nointer_relab, part="dist", vr.name="fcal"))
res_relab$bugs <- rownames(res_relab)
res_relab$method <- "ZIGMM"

res_total <- rbind(res, res_relab)
res_total$p_adj <- p.adjust(res_total$pvalue, method="BH")

write.table(res_total, file = paste0(path, "Results/", "Paraclinical_disease.txt"), col.names=T, row.names=F)
res_disease <- res_total

#Result: Disease score without interaction, zero-inflated:
res=as.data.frame(get.fixed(f_nointer, part="zero", vr.name="fcal"))
res$bugs <- rownames(res)
res$method <- "NBZIMM"

res_relab=as.data.frame(get.fixed(f_nointer_relab, part="zero", vr.name="fcal"))
res_relab$bugs <- rownames(res_relab)
res_relab$method <- "ZIGMM"

res_total <- rbind(res, res_relab)
res_total$p_adj <- p.adjust(res_total$pvalue, method="BH")

write.table(res_total, file = paste0(path, "Results/", "Paraclinical_zero_disease.txt"), col.names=T, row.names=F)
res_disease_zero <- res_total

```

# Test the association with paraclinical disease score using clr trasformation and ZIGMM

Perform the same analyses but with clr transformed microbiome data, to compare the two methods:
```{r}
#Different clr transformation: mclr. It keeps the zeroes as they are, calculate clr from the nonzero data and shifts the clr transformed data to positive values. 
#Analysis of MetaPhlAn count data - all taxa!:
pheno_clr <-  as.matrix(t(otu_table(ps_count)))
pheno_clr <- mclr(pheno_clr)

#Fit using ZIGMM with interaction term:
f_inter_clr <- mms(y=pheno_clr, fixed= ~study_gr+sex+asa+bio+pred+age_gr*fcal, data=clinical, random = ~1|person_ID, method="zig", correlation=corAR1(form=~time_point|person_ID), zi_fixed=~study_gr+sex+asa+bio+pred+age_gr*fcal)

#Find taxa with significant (p-adj<0.05) interaction.  
res_inter_clr=as.data.frame(get.fixed(f_inter_clr, part="dist", vr.name="age_gr2:fcal"))
res_inter_clr$bugs <- rownames(res_inter_clr)
res_inter_clr$method <- "CLR-ZIGMM"
res_inter_clr$p_adj <- p.adjust(res_inter_clr$pvalue, method = "BH")
write.table(res_inter_clr, file = paste0(path, "Results/", "CLR_Paraclinical_interaction.txt"), col.names=T, row.names=F)

res_inter_clr_sig <- res_inter_clr %>% filter(p_adj < 0.05)

##Select the rest of those not significant and fit without the interaction term:
pheno_clr_1 <- pheno_clr[, setdiff(colnames(pheno_clr), rownames(res_inter_clr_sig))]

f_nointer_clr<- mms(y=pheno_clr_1, fixed= ~study_gr+sex+asa+bio+pred+age_gr+fcal, data=clinical, random = ~1|person_ID, method="zig", correlation=corAR1(form=~time_point|person_ID), zi_fixed=~study_gr+sex+asa+bio+pred+age_gr+fcal)


#Result: Disease score in interaction:
res=as.data.frame(get.fixed(f_inter_clr, part="dist", vr.name="fcal"))
res$bugs <- rownames(res)
res$method <- "CLR-ZIGMM"
res$p_adj <- p.adjust(res$pvalue, method="BH")

write.table(res, file = paste0(path, "Results/", "CLR_Paraclinical_interaction_disease.txt"), col.names=T, row.names=F)
res_inter_disease_clr <- res

#Result: Interaction in zero-inflated part:
res=as.data.frame(get.fixed(f_inter_clr, part="zero", vr.name="age_gr2:fcal"))
res$bugs <- rownames(res)
res$method <- "CLR-ZIGMM"
res$p_adj <- p.adjust(res$pvalue, method="BH")

write.table(res, file = paste0(path, "Results/", "CLR_Paraclinical_interaction_zero.txt"), col.names=T, row.names=F)
res_inter_zero_clr <- res

#Result: Disease score in interaction, zero-inflated:
res=as.data.frame(get.fixed(f_inter_clr, part="zero", vr.name="fcal"))
res$bugs <- rownames(res)
res$method <- "CLR-ZIGMM"
res$p_adj <- p.adjust(res$pvalue, method="BH")

write.table(res, file = paste0(path, "Results/", "CLR_Paraclinical_interaction_zero_disease.txt"), col.names=T, row.names=F)
res_inter_zero_disease_clr <- res

#Result: Disease score without interaction:
res=as.data.frame(get.fixed(f_nointer_clr, part="dist", vr.name="fcal"))
res$bugs <- rownames(res)
res$method <- "CLR-ZIGMM"
res$p_adj <- p.adjust(res$pvalue, method="BH")

write.table(res, file = paste0(path, "Results/", "CLR_Paraclinical_disease.txt"), col.names=T, row.names=F)
res_disease_clr <- res

#Result: Disease score without interaction, zero-inflated:
res=as.data.frame(get.fixed(f_nointer_clr, part="zero", vr.name="fcal"))
res$bugs <- rownames(res)
res$method <- "CLR-ZIGMM"
res$p_adj <- p.adjust(res$pvalue, method="BH")

write.table(res, file = paste0(path, "Results/", "CLR_Paraclinical_zero_disease.txt"), col.names=T, row.names=F)
res_disease_zero_clr <- res
```

# Evaluate results from paraclinical disease score association analyses

Investigate the results from the count and relab data: 
```{r}
#Make histograms of p-values from the interaction analysis:
ggplot(res_inter)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Interaction term")


#Extract the significant in interaction term together with the disease score estimate
sig <- res_inter %>% filter(p_adj<0.05)
sig
sig1 <- res_inter_disease[row.names(sig),]
sig1
sig_inter <- sig

#Plot significant interaction:
#Plot the detected species - interaction:
pheno_temp <- pheno
pheno_temp$J_ID <- row.names(pheno_temp)
temp_meta <- merge(pheno_temp, meta, by.x="J_ID", by.y="J_ID")
temp_meta$age_gr1 <- ifelse(temp_meta$age_gr==1, "Children", "Adults")
temp_meta$age_gr1 <-factor(temp_meta$age_gr1, levels=c("Children", "Adults"))

spec <- sig$bugs
for (i in spec){
  j <- gsub("s__", "", i)
  j <- gsub("_", " ", j)
  p <- ggplot(data=temp_meta, aes_string(x="f_cal_current", y=paste0(i, "/rdepth"), color = "age_gr1"))+
    geom_point(size=1)+geom_smooth(se=F)+theme_classic()+xlab("Faecal calprotectin")+ylab("Relative abundance") + ggtitle(as.character(j))+labs(color="Age group")
print(p)

}


#Extract the significant in interaction term in the zero-inflated part, together with the disease score estimate
ggplot(res_inter_zero)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Zero-inflated interaction term")

sig <- res_inter_zero %>% filter(p_adj<0.05)
sig 
sig1 <- res_inter_zero_disease[row.names(sig),]
sig1
sig_inter_zero <- sig

spec <- sig$bugs
for (i in spec){
  test <- temp_meta[,i]
  test <- ifelse(test==0, "Absent", "Present")
  temp_meta$test <- test
  j <- gsub("s__", "", i)
  j <- gsub("_", " ", j)
  p <- ggplot(data=temp_meta)+
    geom_boxplot(aes(x=test, y=f_cal_current, color=test), outlier.size=1)+theme_classic()+xlab(" ")+ylab("Faecal calprotectin") + ggtitle(as.character(j))+theme(legend.position="none")+facet_wrap(~age_gr1)
 print(p)

}

#Plot all NBZIMM interaction results together in one plot:
sig <- res_inter %>% filter(p_adj<0.05)
spec <- c(sig$bugs, "s__Eubacterium_siraeum")
pheno_temp_relab <- pheno_relab
pheno_temp_relab$J_ID <- row.names(pheno_temp_relab)
temp_meta_relab <- merge(pheno_temp_relab, meta, by.x="J_ID", by.y="J_ID")
temp_meta_relab$age_gr1 <- ifelse(temp_meta_relab$age_gr==1, "Children", "Adults")
temp_meta_relab$age_gr1 <-factor(temp_meta_relab$age_gr1, levels=c("Children", "Adults"))
temp1 <- temp_meta_relab %>%
  pivot_longer(., cols = colnames(pheno_relab), names_to = "Var", values_to = "Val")

temp1$Var1 <- gsub("s__", "", temp1$Var)
temp1$Var1 <- gsub("_", " ", temp1$Var1)

temp2 <- temp1 %>% filter(Var %in% spec)
ggplot(temp2, aes(x=f_cal_current, y=Val, color=age_gr1))+
  geom_point(size=1)+geom_smooth(se=F)+theme_classic()+facet_wrap(~Var1, scales="free_y")+ylab("Relative abundance")+labs(color="Age group")+ theme(legend.position = c(1, 0),legend.justification = c(1, 0))+xlab("Faecal calprotectin")


#Investigate results without interaction term: Disease score:
ggplot(res_disease)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Disease score")

sig <- res_disease %>% filter(p_adj<0.05)
sig
sig_disease <- sig

#Plot the detected species:
spec <- sig$bugs
for (i in spec){
  j <- gsub("s__", "", i)
  j <- gsub("_", " ", j)
  p <- ggplot(data=temp_meta, aes_string(x="f_cal_current", y=paste0(i, "/rdepth")))+
    geom_point(size=1)+geom_smooth(se=T)+theme_classic()+xlab("Faecal calprotectin")+ylab("Relative abundance") + ggtitle(as.character(j))
 print(p)

}

#Plot all together:
temp2 <- temp1 %>% filter(Var %in% spec)
ggplot(temp2, aes(x=f_cal_current, y=Val))+
  geom_point(size=1)+geom_smooth(se=T)+theme_classic()+facet_wrap(~Var1, scales="free_y", ncol=3)+ylab("Relative abundance")+xlab("Faecal calprotectin")

#Investigate results without interaction term: Disease score, zero inflated:
ggplot(res_disease_zero)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Zero-inflated disease score")

sig <- res_disease_zero %>% filter(p_adj<0.05)
sig #None

```

Visualize the CLR results and compare!
```{r}
#Make histograms of p-values from the interaction analysis:
ggplot(res_inter_clr)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Interaction term")


#Extract the significant in interaction term together with the disease score estimate
sig <- res_inter_clr %>% filter(p_adj<0.05)
sig
sig1 <- res_inter_disease_clr[row.names(sig),]
sig1
sig_inter_clr <- sig

spec <- sig$bugs
for (i in spec){
  p <- ggplot(data=temp_meta, aes_string(x="f_cal_current", y=paste0(i, "/rdepth"), color = "age_gr"))+
    geom_point()+geom_smooth(se=F)+theme_classic()+xlab("Faecal calprotectin")+ylab("Relative abundance") + ggtitle(as.character(i))+labs(color="Age group")
  print(p)
}



#Extract the significant in interaction term in the zero-inflated part, together with the disease score estimate
ggplot(res_inter_zero_clr)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Zero-inflated interaction term")

sig <- res_inter_zero_clr %>% filter(p_adj<0.05)
sig
sig1 <- res_inter_zero_disease_clr[row.names(sig),]
sig1
sig_inter_zero_clr <- sig

#Plot the results:
#Plot the detected species:
spec <- sig$bugs
for (i in spec){
  test <- temp_meta[,i]
  test <- ifelse(test==0, "Absent", "Present")
  temp_meta$test <- test
  p <- ggplot(data=temp_meta)+
    geom_boxplot(aes(x=test, y=f_cal_current, color=age_gr))+theme_classic()+xlab(" ")+ylab("Faecal calprotectin") + ggtitle(as.character(i))+labs(color="Age group")
  print(p)

}

#Investigate results without interaction term: Disease score:
ggplot(res_disease_clr)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Disease score")

sig <- res_disease_clr %>% filter(p_adj<0.05)
sig
sig_disease_clr <- sig

#Plot the detected species:
spec <- sig$bugs
for (i in spec){
  p <- ggplot(data=temp_meta)+
    geom_point(aes_string(x="f_cal_current", y=paste0(i, "/rdepth")))+theme_classic()+xlab("Faecal calprotectin")+ylab("Relative abundance") + ggtitle(as.character(i))+theme(legend.position = "none")
 print(p)

}


#Investigate results without interaction term: Disease score, zero inflated:
ggplot(res_disease_zero_clr)+
  geom_histogram(aes(x=pvalue), bins=30, breaks = seq(0, 1, by=1/30), color="black", fill = "lightblue")+
  theme_classic()+ xlab("p-value")+ylab("Count")+ggtitle("Zero-inflated disease score")

sig <- res_disease_zero_clr %>% filter(p_adj<0.05)
sig #None!
```

Compare the two methods:
```{r}
sig_inter$test <- "Interaction: NBZIMM, NB"
sig_inter_zero$test <- "Interaction: NBZIMM, ZI"
sig_inter_clr$test <- "Interaction: ZIGMM, G"
sig_inter_zero_clr$test <- "Interaction: ZIGMM, ZI"
sig_disease$test <- "Disease score: NBZIMM, NB"
sig_disease_clr$test <- "Disease score: ZIGMM, G"

#Change to enable comparison:Transformed to model presence of species instead of absence
sig_inter_zero_clr$Estimate <- -sig_inter_zero_clr$Estimate
sig_inter_zero$Estimate <- -sig_inter_zero$Estimate


res_total <- rbind(sig_inter, sig_inter_zero, sig_inter_clr, sig_inter_zero_clr, sig_disease, sig_disease_clr)
res_total$test <- factor(res_total$test, levels=c("Interaction: NBZIMM, NB", "Interaction: NBZIMM, ZI","Interaction: ZIGMM, G", "Interaction: ZIGMM, ZI", "Disease score: NBZIMM, NB","Disease score: ZIGMM, G"))

bl <- colorRampPalette(c("navy","royalblue","lightskyblue"))(200)   
re <- colorRampPalette(c("mistyrose", "red2","darkred"))(200)

min(res_total$Estimate)
max(res_total$Estimate)

res_total$bugs <- gsub("s__", "", res_total$bugs)
res_total$bugs <- gsub("_", " ", res_total$bugs)

ggplot(data=res_total)+
  geom_tile(aes(x=test, y=bugs, fill=Estimate))+ylab("")+xlab(" ")+
  scale_fill_gradientn(colours=c(bl, re), na.value = "black",
         limits = c(-1.2, 1.2))+theme_classic()+theme(axis.text.x = element_text(angle = 90))

```
