---
title: "Investigation of diversity and stability"
output:
  github_document:
    toc: true
editor_options:
  chunk_output_type: console
---
# Read in packages and data

Read in packages and path to folder:
```{r echo=T, message=F, error=F, warning=F}
library(dplyr)
library(ggplot2)
library(readr)
library(ggpubr)
library(vegan)
library(caret)
library(phyloseq)
library(lme4)
library(lmerTest)
library(nlme)
library(RCurl)
library(MuMIn)
main.dir = "C:/Users/FX76TZ/OneDrive - Aalborg Universitet/IBD project/UC - biopsier/"
setwd(paste0( file.path( main.dir, "data/16S/data_processed")))
```

Read in data:
```{r echo=T, message=F, error=F, warning=F}
#Phyloseq objects:
ps_relab = readRDS("DATA.ps.raw_SampleFilter.relab_meta.modified.rds")
ps_count = readRDS( "DATA.ps.raw_SampleFilter_meta.modified.rds")
meta <- as.data.frame(as.matrix(sample_data(ps_relab)))
meta$J_ID <- row.names(meta)

sample_data(ps_relab)$GI.location <- ordered(sample_data(ps_relab)$GI.location, levels = c('r','s',  't', 'c'), labels = c('r','s',  't', 'c'))
sample_data(ps_count)$GI.location <- ordered(sample_data(ps_count)$GI.location, levels = c('r','s',  't', 'c'), labels = c('r','s',  't', 'c'))

sample_data(ps_relab)$person_ID <- sample_data(ps_relab)$study_id_new_meta
sample_data(ps_relab)$age_gr <- as.factor(ifelse(sample_data(ps_relab)$age_gr=="Paediatric", 1, ifelse(sample_data(ps_relab)$age_gr=="Adult", 2, NA)))
sample_data(ps_count)$person_ID <- sample_data(ps_count)$study_id_new_meta
sample_data(ps_count)$age_gr <- as.factor(ifelse(sample_data(ps_count)$age_gr=="Paediatric", 1, ifelse(sample_data(ps_count)$age_gr=="Adult", 2, NA)))

sample_data(ps_relab)$J_ID <- row.names(sample_data(ps_relab))
sample_data(ps_count)$J_ID <- row.names(sample_data(ps_count))

#Make alpha diversity and merge with meta data:
matrix_spec <- t(otu_table(ps_relab))
div_shannon_003 <- diversity(t(matrix_spec), index="shannon")  ### Shannon entropy
spec_nr_003 <- specnumber(matrix_spec,MARGIN=2)             ### Species number (observed)

alpha_div_relab <- as.data.frame(cbind(spec_nr_003,div_shannon_003))
alpha_div_relab$J_ID <- row.names(alpha_div_relab)
alpha_div <- merge(alpha_div_relab, meta, by.x="J_ID", by.y="J_ID")
alpha_div$GI.location <- ordered(alpha_div$GI.location, levels = c('r','s',  't', 'c'), labels = c('r','s',  't', 'c'))
head(alpha_div_relab)
```

Investigate if alpha diversity and unmapped reads are normally distributed and transform if not!:
```{r}
#Chao1
ggplot(data=alpha_div)+
  geom_histogram(aes(x=spec_nr_003), color="black", fill = "lightblue", bins=40)+theme_classic()+xlab("Observed diveristy")+ylab("Count")

#Shannon
ggplot(data=alpha_div)+
  geom_histogram(aes(x=div_shannon_003), color="black", fill = "lightblue", bins=40)+theme_classic()+xlab("Shannon diversity")+ylab("Count")

alpha_div$GI.location <- ifelse(alpha_div$GI.location == "c", "Caecum", ifelse(alpha_div$GI.location == "r", "Rectum", ifelse(alpha_div$GI.location == "s", "Sigmoid colon", "Tansverse colon")))

ggplot(data=alpha_div)+
  geom_boxplot(aes(x=GI.location, y=spec_nr_003))+xlab("")+ylab("Observed richness")+theme_classic()

ggplot(data=alpha_div)+
  geom_boxplot(aes(x=GI.location, y=div_shannon_003))+xlab("")+ylab("Shannon diversity")+theme_classic()
```

Test: Does alpha diversity associate with age, study group and/or disease score?
Tested using linear mixed models
```{r warning=FALSE}
alpha_div$person_ID <- alpha_div$study_id_new_meta
alpha_div$age_gr <- as.factor(ifelse(alpha_div$age_gr=="Paediatric", 1, ifelse(alpha_div$age_gr=="Adult", 2, NA)))
#Discrete numeric disease score
divs <- c("spec_nr_003", "shannon")

#Within study groups: Does alpha associate with age group?:
alpha_new <- alpha_div %>% filter(study_gr==1)
alpha_old <- alpha_div %>% filter(study_gr==2)

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr"))
  model1 <- lme(form, data=alpha_new, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ study_gr+age_gr"))
  model1 <- lme(form, data=alpha_div, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

#What is the association with disease score? Include age:
#Should interaction with age be included?
for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+as.numeric(geb)+as.numeric(geb)*age_gr"))
  model1 <- lme(form, data=alpha_new, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+as.numeric(geb)+as.numeric(geb)*age_gr"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}
for (div in divs){
  form <- as.formula(paste0(div, "~ study_gr+age_gr+as.numeric(geb)+as.numeric(geb)*age_gr"))
  model1 <- lme(form, data=alpha_div, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

#Interaction is not significant. Fit without:

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+as.numeric(geb)"))
  model1 <- lme(form, data=alpha_new, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+as.numeric(geb)"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}
for (div in divs){
  form <- as.formula(paste0(div, "~ study_gr+age_gr+as.numeric(geb)"))
  model1 <- lme(form, data=alpha_div, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}


#Split the dataset by age group and test influence of study group:
alpha_children <- alpha_div %>% filter(age_gr==1)
alpha_adults <- alpha_div %>% filter(age_gr==2)

for (div in divs){
  form <- as.formula(paste0(div, "~ study_gr"))
  model1 <- lme(form, data=alpha_children, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ study_gr"))
  model1 <- lme(form, data=alpha_adults, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+study_gr"))
  model1 <- lme(form, data=alpha_div, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

#Influence of medicine for earlier diagnosed:
for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+bio_0"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+aza_0"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+pred_0"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+asa_0"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

#Also include geb:
for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+as.numeric(geb)+bio_0"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+as.numeric(geb)+aza_0"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+as.numeric(geb)+pred_0"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+as.numeric(geb)+asa_0"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}


#Test difference between GI locations: 
for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+GI.location"))
  model1 <- lme(form, data=alpha_new, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+GI.location"))
  model1 <- lme(form, data=alpha_old, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}

for (div in divs){
  form <- as.formula(paste0(div, "~ age_gr+study_gr+GI.location"))
  model1 <- lme(form, data=alpha_div, random=~1|person_ID, na.action=na.omit, correlation=corLin(form=~GI.location|person_ID))
  print(summary(model1)$tTable)
}



```

# Does beta diversity associate with age, study group and/or disease score?
Tested with PERMANOVA
```{r}
#Number of samples/participant must be even! Select samples with 2 data points
ent <- subset_samples(ps_relab, !is.na(geb))
df <- data.frame(sample_data(ent))
n_df <- df %>% group_by(person_ID) %>% summarise(n=n()) %>% filter(n>1)
df_reduced <- df %>% filter(person_ID %in% n_df$person_ID)
reduce <- df_reduced %>% group_by(person_ID)%>% arrange(GI.location) %>% slice(1:2) %>% ungroup()
ps_reduced <- subset_samples(ps_relab, J_ID %in% reduce$J_ID)

#Divide by study group:
ps_new <- subset_samples(ps_reduced, study_gr==1)
ps_old <- subset_samples(ps_reduced, study_gr==2)

#Test age group - newly diagnosed:
perm <- how(within=Within(type="none"), plots=Plots(strata=sample_data(ps_new)$person_ID, type="free"), nperm=999)

set.seed(1)
adonis2(otu_table(ps_new)~age_gr, data=as.data.frame(as.matrix(sample_data(ps_new))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_new)~age_gr, data=as.data.frame(as.matrix(sample_data(ps_new))), method="jaccard", permutations=perm, by="margin", binary=T)

#Test age group - previously diagnosed:
perm <- how(within=Within(type="none"), plots=Plots(strata=sample_data(ps_old)$person_ID, type="free"), nperm=999)

set.seed(1)
adonis2(otu_table(ps_old)~age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_old)~age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="jaccard", permutations=perm, by="margin", binary=T)

#Test age group - full dataset:
perm <- how(within=Within(type="none"), plots=Plots(strata=sample_data(ps_reduced)$person_ID, type="free"), nperm=999)

set.seed(1)
adonis2(otu_table(ps_reduced)~age_gr+study_gr, data=as.data.frame(as.matrix(sample_data(ps_reduced))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_reduced)~age_gr+study_gr, data=as.data.frame(as.matrix(sample_data(ps_reduced))), method="jaccard", permutations=perm, by="margin", binary=T)


#Test the influence of f-cal:
ps_reduced <- subset_samples(ps_reduced, !is.na(geb))
ps_new <- subset_samples(ps_reduced, study_gr==1)
ps_old <- subset_samples(ps_reduced, study_gr==2)

#Newly diagnosed:
perm <- how(within=Within(type="free"), plots=Plots(strata=sample_data(ps_new)$person_ID, type="none"), nperm=999)

set.seed(1)
adonis2(otu_table(ps_new)~as.numeric(geb)+age_gr, data=as.data.frame(as.matrix(sample_data(ps_new))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_new)~as.numeric(geb)+age_gr, data=as.data.frame(as.matrix(sample_data(ps_new))), method="jaccard", permutations=perm, by="margin", binary=T)

#Previously diagnosed:
perm <- how(within=Within(type="free"), plots=Plots(strata=sample_data(ps_old)$person_ID, type="none"), nperm=999)

set.seed(1)
adonis2(otu_table(ps_old)~as.numeric(geb)+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_old)~as.numeric(geb)+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="jaccard", permutations=perm, by="margin", binary=T)

#Full dataset:
perm <- how(within=Within(type="free"), plots=Plots(strata=sample_data(ps_reduced)$person_ID, type="none"), nperm=999)

set.seed(1)
adonis2(otu_table(ps_reduced)~as.numeric(geb)+age_gr+study_gr, data=as.data.frame(as.matrix(sample_data(ps_reduced))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_reduced)~as.numeric(geb)+age_gr+study_gr, data=as.data.frame(as.matrix(sample_data(ps_reduced))), method="jaccard", permutations=perm, by="margin", binary=T)


# Divide by age group and test study group:
ps_children <- subset_samples(ps_reduced, age_gr==1)
ps_adults <- subset_samples(ps_reduced, age_gr==2)

#Children:
perm <- how(within=Within(type="none"), plots=Plots(strata=sample_data(ps_children)$person_ID, type="free"), nperm=999)

set.seed(1)
adonis2(otu_table(ps_children)~study_gr, data=as.data.frame(as.matrix(sample_data(ps_children))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_children)~study_gr, data=as.data.frame(as.matrix(sample_data(ps_children))), method="jaccard", permutations=perm, by="margin", binary=T)

#Adults:
perm <- how(within=Within(type="none"), plots=Plots(strata=sample_data(ps_adults)$person_ID, type="free"), nperm=999)

set.seed(1)
adonis2(otu_table(ps_adults)~study_gr, data=as.data.frame(as.matrix(sample_data(ps_adults))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_adults)~study_gr, data=as.data.frame(as.matrix(sample_data(ps_adults))), method="jaccard", permutations=perm, by="margin", binary=T)

#Full dataset:
perm <- how(within=Within(type="none"), plots=Plots(strata=sample_data(ps_reduced)$person_ID, type="free"), nperm=999)

set.seed(1)
adonis2(otu_table(ps_reduced)~age_gr+study_gr, data=as.data.frame(as.matrix(sample_data(ps_reduced))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_reduced)~age_gr+study_gr, data=as.data.frame(as.matrix(sample_data(ps_reduced))), method="jaccard", permutations=perm, by="margin", binary=T)


#Determine the effect of medicine for patients with older diagnosis:
perm <- how(within=Within(type="none"), plots=Plots(strata=sample_data(ps_old)$person_ID, type="free"), nperm=999)

set.seed(1)
adonis2(otu_table(ps_old)~bio_0+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_old)~bio_0+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="jaccard", permutations=perm, by="margin", binary=T)

set.seed(1)
adonis2(otu_table(ps_old)~aza_0+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_old)~aza_0+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="jaccard", permutations=perm, by="margin", binary=T)

set.seed(1)
adonis2(otu_table(ps_old)~pred_0+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_old)~pred_0+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="jaccard", permutations=perm, by="margin", binary=T)

set.seed(1)
adonis2(otu_table(ps_old)~asa_0+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_old)~asa_0+age_gr, data=as.data.frame(as.matrix(sample_data(ps_old))), method="jaccard", permutations=perm, by="margin", binary=T)


#Test the influence of GI-segments:
#Newly diagnosed:
perm <- how(within=Within(type="free"), plots=Plots(strata=sample_data(ps_new)$person_ID, type="none"), nperm=999)

set.seed(1)
adonis2(otu_table(ps_new)~age_gr+GI.location, data=as.data.frame(as.matrix(sample_data(ps_new))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_new)~age_gr+GI.location, data=as.data.frame(as.matrix(sample_data(ps_new))), method="jaccard", permutations=perm, by="margin", binary=T)

#Previously diagnosed:
perm <- how(within=Within(type="free"), plots=Plots(strata=sample_data(ps_old)$person_ID, type="none"), nperm=999)

set.seed(1)
adonis2(otu_table(ps_old)~age_gr+GI.location, data=as.data.frame(as.matrix(sample_data(ps_old))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_old)~age_gr+GI.location, data=as.data.frame(as.matrix(sample_data(ps_old))), method="jaccard", permutations=perm, by="margin", binary=T)

#Full dataset:
perm <- how(within=Within(type="free"), plots=Plots(strata=sample_data(ps_reduced)$person_ID, type="none"), nperm=999)

set.seed(1)
adonis2(otu_table(ps_reduced)~age_gr+GI.location+study_gr, data=as.data.frame(as.matrix(sample_data(ps_reduced))), method="bray", permutations=perm, by="margin")

set.seed(1)
adonis2(otu_table(ps_reduced)~age_gr+GI.location+study_gr, data=as.data.frame(as.matrix(sample_data(ps_reduced))), method="jaccard", permutations=perm, by="margin", binary=T)
```
